---
title: "01_dependencies"
format: pdf
editor: visual
---

Installation and loading of all packages required by the pipeline. Packages are installed into a project-local library at `dependencies/r_packages/` to keep the environment self-contained.

Exact package versions used in the study are recorded in `dependencies/package_versions.csv`.

```{r}
packages <- read.csv(here::here("dependencies/package_versions.csv"))
# Install each package at the recorded version via remotes
remotes::install_version(packages$Package[i], version = packages$Version[i])
```

Subsequent scripts load packages from this local library via the `.libPaths()` call in the "Add library to path" section below.

------------------------------------------------------------------------

# System dependencies

The following system libraries are required before installing R packages. Run these commands in a terminal (requires sudo on Linux):

```{bash}
#| eval: false
sudo apt-get update
sudo apt-get install -y \
  bedtools \
  libhdf5-dev \
  libhdf5-serial-dev \
  build-essential \
  zlib1g-dev \
  libssl-dev \
  libssh2-1-dev \
  libcurl4-openssl-dev \
  r-base-dev \
  liblzma-dev \
  libbz2-dev
```

------------------------------------------------------------------------

# Create project-local R library

```{r}
pkg_path <- here::here("dependencies/r_packages")
dir.create(pkg_path, recursive = TRUE, showWarnings = FALSE)
```

------------------------------------------------------------------------

# Install R packages

```{r}
#| eval: false
#| output: false

# CRAN
install.packages(c(
  "BiocManager",
  "devtools",
  "remotes",
  "pheatmap",
  "qlcMatrix",
  "aricode",
  "harmony",
  "tictoc",
  "hdf5r",
  "patchwork",
  "distances",
  "spdep",
  "pcaPP",
  "clustree",
  "ggridges",
  "viridis",
  "tidyverse",
  "tidyr",
  "dplyr",
  "ggplot2"
), lib = pkg_path)

# Bioconductor
BiocManager::install(c(
  "BiocParallel",
  "BiocNeighbors",
  "DropletUtils",
  "Seurat",
  "celda",
  "HGC",
  "SingleR",
  "UCell",
  "scuttle",
  "scran",
  "scater",
  "glmGamPoi",
  "decontX",
  "writexl",
  "orthogene",
  "org.Hs.eg.db"
), lib = pkg_path)

# GitHub
devtools::install_github("satijalab/seurat-data",      lib = pkg_path)
devtools::install_github("LungCellAtlas/FastCAR",       lib = pkg_path)
devtools::install_github("ZJU-UoE-CCW-LAB/scCDC",      lib = pkg_path)
devtools::install_github("wjawaid/enrichR",             lib = pkg_path)
devtools::install_github("MarioniLab/miloR",            lib = pkg_path)
devtools::install_github("MarioniLab/miloDE",           lib = pkg_path)
devtools::install_github("theislab/kBET",               lib = pkg_path)
devtools::install_github("immunogenomics/presto",       lib = pkg_path)
devtools::install_github("immunogenomics/lisi",         lib = pkg_path)
remotes::install_github("satijalab/azimuth", ref = "master", lib = pkg_path)

install.packages("SoupX",    lib = pkg_path)
install.packages("ggrastr",  lib = pkg_path)
install.packages("cluster",  lib = pkg_path)
install.packages("DescTools", lib = pkg_path)
install.packages("plyr",     lib = pkg_path)
```

------------------------------------------------------------------------

# Save package versions

```{r}
installed <- as.data.frame(installed.packages(lib.loc = here::here("dependencies/r_packages")))
write.csv(installed[, c("Package", "Version")], 
          here::here("dependencies/package_versions.csv"), 
          row.names = FALSE)
```

------------------------------------------------------------------------

# Add library to path

Add the project-local library to the search path. This line should be run at the start of every session (it is called automatically by sourcing `scripts/R/R_files/01_functions.R`).

```{r}
.libPaths(c(here::here("dependencies/r_packages"), .libPaths()))
```

------------------------------------------------------------------------

# Load libraries

```{r}
#| message: false
#| warning: false
library(here)
library(Matrix)
library(Seurat)
library(SeuratObject)
library(SeuratData)
library(FastCAR)
library(scCDC)
library(SoupX)
library(ComplexHeatmap)
library(circlize)
library(HGC)
library(aricode)
library(lisi)
library(SingleR)
library(scuttle)
library(BiocParallel)
library(BiocNeighbors)
library(scran)
library(scater)
library(qlcMatrix)
library(viridis)
library(DropletUtils)
library(patchwork)
library(harmony)
library(ggridges)
library(decontX)
library(tictoc)
library(glmGamPoi)
library(hdf5r)
library(rhdf5)
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
library(presto)
library(miloDE)
library(miloR)
library(distances)
library(cluster)
library(Azimuth)
library(kBET)
library(DescTools)
library(plyr)
library(writexl)
library(ggrastr)
library(clustree)
library(orthogene)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(spdep)
library(pcaPP)
library(UCell)
```
