---
title: "02_functions"
format: pdf
editor: visual
---

All self-written functions used throughout this study are listed here, organised by the analysis step in which they are used.

**`scripts/R/R_files/01_functions.R` is the canonical source of all functions.** This `.qmd` file documents those functions with formatted prose and roxygen headers, but the code shown here mirrors `01_functions.R` exactly. Downstream scripts source the `.R` file directly.

# Source functions

```{r}
# All downstream scripts begin with this line
source(here::here("scripts/R/R_files/01_functions.R"))
```

------------------------------------------------------------------------

# Preprocessing and quality control

```{r}
#' Load a STARsolo / CellRanger output folder
#'
#' Reads barcodes.tsv, features.tsv, and matrix.mtx from a directory and
#' returns a sparse count matrix.
#'
#' @param folder Character. Path to the folder containing the three files.
#'
#' @return A sparse CsparseMatrix (genes x barcodes).
load_folder <- function(folder) {
  barcodes <- read.delim(file.path(folder, "barcodes.tsv"), header = FALSE)
  features <- read.delim(file.path(folder, "features.tsv"), header = FALSE)
  counts   <- Matrix::readMM(file.path(folder, "matrix.mtx"))
  colnames(counts) <- barcodes$V1
  rownames(counts) <- features$V2
  counts <- as(counts, "CsparseMatrix")
  return(counts)
}
```

------------------------------------------------------------------------

# Running decontamination methods

## R-based methods

```{r}
#' Run all R-based ambient RNA decontamination methods
#'
#' Runs FastCAR, scCDC, DecontX (full and reduced), and SoupX (full and
#' reduced) on one or more samples. Clustering is performed with Seurat
#' (and Harmony for multi-sample datasets) before the methods are applied.
#'
#' @param counts Named list of raw count matrices (genes x barcodes), one
#'   per sample.
#' @param barcodes Named list of character vectors giving the cell barcodes
#'   to retain (e.g. post-emptyDrops), one per sample.
#' @param dims Integer. PCA / Harmony dimensions for neighbour finding
#'   (default 10).
#' @param resolution Numeric. Seurat clustering resolution (default 0.1).
#' @param nfeat Integer. Number of variable features (default 2000).
#'
#' @return A Seurat object containing the RNA assay plus one assay per
#'   method: FastCAR, scCDC, SoupXfull, SoupXreduced, DecontXfull,
#'   DecontXreduced.
run_decontamination_methods <- function(counts,
                                        barcodes,
                                        dims       = 10,
                                        resolution = 0.1,
                                        nfeat      = 2000) {

  # ------------------------------------------------------------------
  # 1. Subset matrices and build per-sample Seurat objects
  # ------------------------------------------------------------------
  message("Setting up data and clustering.")

  seu_list    <- list()
  subset_list <- list()

  for (dataset in seq_along(counts)) {

    # Append sample name to barcodes to ensure uniqueness across samples
    colnames(counts[[dataset]]) <- paste(
      colnames(counts[[dataset]]), names(counts)[dataset], sep = "_"
    )
    barcodes[[dataset]] <- paste(
      barcodes[[dataset]], names(counts)[dataset], sep = "_"
    )

    # Subset to called cells; remove all-zero genes
    subset_list[[dataset]] <- counts[[dataset]][
      , colnames(counts[[dataset]]) %in% barcodes[[dataset]]
    ]
    subset_list[[dataset]] <- subset_list[[dataset]][
      Matrix::rowSums(subset_list[[dataset]]) >= 1, 
    ]
    names(subset_list)[dataset]          <- names(counts)[dataset]
    rownames(subset_list[[dataset]])     <- make.unique(rownames(subset_list[[dataset]]))

    # Create Seurat object
    seu_list[[dataset]] <- suppressWarnings(
      CreateSeuratObject(
        counts       = subset_list[[dataset]],
        project      = names(counts)[dataset],
        min.cells    = 0,
        min.features = 0
      )
    )

    # Store barcode as metadata for downstream matching
    seu_list[[dataset]]@meta.data$barcode <- colnames(subset_list[[dataset]])

    # Basic QC: >= 200 features per cell; gene detected in >= 3 cells
    seu_list[[dataset]] <- subset(
      seu_list[[dataset]],
      cells = rownames(seu_list[[dataset]]@meta.data)[
        seu_list[[dataset]]$nFeature_RNA >= 200
      ]
    )
    seu_list[[dataset]] <- subset(
      seu_list[[dataset]],
      features = names(which(
        rowSums(seu_list[[dataset]]@assays$RNA$counts > 0) >= 3
      ))
    )
  }

  # ------------------------------------------------------------------
  # 2. Merge samples and cluster
  # ------------------------------------------------------------------
  if (length(counts) > 1) {
    seu <- suppressWarnings(merge(seu_list[[1]], seu_list[2:length(seu_list)]))
  } else {
    seu <- seu_list[[1]]
  }

  seu <- NormalizeData(seu, verbose = FALSE)
  seu <- FindVariableFeatures(seu, verbose = FALSE, nfeatures = nfeat)
  seu <- ScaleData(seu, features = rownames(seu), verbose = FALSE)
  seu <- RunPCA(seu, verbose = FALSE)

  if (length(counts) > 1) {
    seu <- suppressWarnings(
      RunHarmony(seu, group.by.vars = "orig.ident", verbose = FALSE)
    )
    seu <- FindNeighbors(seu, dims = 1:dims, reduction = "harmony", verbose = FALSE)
  } else {
    seu <- FindNeighbors(seu, dims = 1:dims, verbose = FALSE)
  }
  seu <- FindClusters(seu, resolution = resolution, verbose = FALSE)

  # Species fraction metadata (relevant for barnyard / species-mixing datasets)
  seu[["input.percent.mm10"]]   <- PercentageFeatureSet(seu, pattern = "^mm10")
  seu[["input.percent.GRCh38"]] <- PercentageFeatureSet(seu, pattern = "^GRCh38")

  # Re-split merged object into per-sample objects for method execution
  for (dataset in seq_along(counts)) {
    seu_list[[dataset]] <- subset(
      seu,
      cells = rownames(seu@meta.data)[seu$orig.ident == names(counts)[dataset]]
    )
  }

  # ------------------------------------------------------------------
  # 3. Run decontamination methods per sample
  # ------------------------------------------------------------------
  fastcar         <- list()
  cdc_full        <- list()
  soupx_full      <- list()
  soupx_reduced   <- list()
  decontx_full    <- list()
  decontx_reduced <- list()

  for (dataset in seq_along(counts)) {

    # Prepare filtered and raw matrices with matching rownames
    filt_matrix <- seu_list[[dataset]]@assays$RNA$counts
    colnames(filt_matrix) <- seu_list[[dataset]]$barcode

    raw_matrix              <- counts[[dataset]]
    rownames(raw_matrix)    <- gsub("_", "-", rownames(raw_matrix))
    raw_matrix              <- raw_matrix[!duplicated(rownames(raw_matrix)), ]
    filt_matrix             <- filt_matrix[!duplicated(rownames(filt_matrix)), ]
    rownames(filt_matrix)   <- gsub("_", "-", rownames(filt_matrix))
    filt_matrix             <- filt_matrix[rownames(filt_matrix) %in% rownames(raw_matrix), ]
    raw_matrix              <- raw_matrix[rownames(raw_matrix) %in% rownames(filt_matrix), ]
    filt_matrix             <- filt_matrix[match(rownames(raw_matrix), rownames(filt_matrix)), ]

    clusters <- data.frame(
      barcode = seu_list[[dataset]]$barcode,
      cluster = seu_list[[dataset]]$seurat_clusters
    )
    clusters <- clusters[match(colnames(filt_matrix), clusters$barcode), ]

    message(paste0("\nRunning decontamination for: ", names(counts)[dataset]))

    # --- FastCAR ---
    message("\tRunning FastCAR.")
    ambProfile <- FastCAR::describe.ambient.RNA.sequence(
      fullCellMatrix            = counts[[dataset]],
      start                     = 10,
      stop                      = 500,
      by                        = 10,
      contaminationChanceCutoff = 0.05
    )
    emptyDropletCutoff <- FastCAR::recommend.empty.cutoff(ambProfile)
    ambientProfile     <- FastCAR::determine.background.to.remove(
      counts[[dataset]], counts[[dataset]],
      emptyDropletCutoff        = emptyDropletCutoff,
      contaminationChanceCutoff = 0.05
    )
    fastcar[[dataset]] <- FastCAR::remove.background(subset_list[[dataset]], ambientProfile)

    # --- SoupX full (raw + filtered matrices) ---
    message("\tRunning SoupX.")
    sc           <- SoupChannel(raw_matrix, filt_matrix)
    sc           <- setClusters(sc, setNames(clusters$cluster, clusters$barcode))
    sc           <- autoEstCont(sc, doPlot = FALSE, verbose = FALSE)
    soupx_full[[dataset]] <- suppressWarnings(adjustCounts(sc, verbose = FALSE))

    # --- SoupX reduced (filtered matrix only; soup profile from data) ---
    sc_no_drops <- SoupChannel(filt_matrix, filt_matrix, calcSoupProfile = FALSE)
    sc_no_drops <- setClusters(sc_no_drops, setNames(clusters$cluster, clusters$barcode))
    soup_prof   <- data.frame(
      row.names = rownames(filt_matrix),
      est       = rowSums(filt_matrix) / sum(filt_matrix),
      counts    = rowSums(filt_matrix)
    )
    sc_no_drops <- setSoupProfile(sc_no_drops, soup_prof)
    sc_no_drops <- autoEstCont(sc_no_drops, doPlot = FALSE, verbose = FALSE)
    soupx_reduced[[dataset]] <- suppressWarnings(adjustCounts(sc_no_drops, verbose = FALSE))

    # --- scCDC ---
    message("\tRunning scCDC.")
    out <- capture.output(
      contamination <- suppressMessages(suppressWarnings(
        ContaminationDetection(
          seu_list[[dataset]],
          restriction_factor = 0.8,
          sample_name        = names(counts)[dataset]
        )
      ))
    )
    out <- capture.output(
      cdc_full[[dataset]] <- suppressMessages(suppressWarnings(
        ContaminationCorrection(
          seu_list[[dataset]],
          cont_genes = rownames(contamination),
          auc_thres  = 0.9
        )
      ))
    )
    cdc_full[[dataset]]              <- cdc_full[[dataset]]@assays$Corrected$counts
    colnames(cdc_full[[dataset]])    <- seu_list[[dataset]]$barcode

    # --- DecontX full (uses raw matrix as background) ---
    message("\tRunning DecontX.")
    out <- capture.output(
      decontx_full[[dataset]] <- decontX(
        x          = filt_matrix,
        z          = clusters$cluster,
        background = raw_matrix,
        verbose    = FALSE
      )
    )
    decontx_full[[dataset]] <- decontx_full[[dataset]]$decontXcounts

    # --- DecontX reduced (no background matrix) ---
    out <- capture.output(
      decontx_reduced[[dataset]] <- decontX(
        x       = filt_matrix,
        z       = clusters$cluster,
        verbose = FALSE
      )
    )
    decontx_reduced[[dataset]] <- decontx_reduced[[dataset]]$decontXcounts
  }

  # ------------------------------------------------------------------
  # 4. Add corrected counts as assays to the merged Seurat object
  # ------------------------------------------------------------------
  message("\nAdding assays.")
  seu <- add_assay(seu, fastcar,         "FastCAR")
  seu <- add_assay(seu, cdc_full,        "scCDC")
  seu <- add_assay(seu, soupx_full,      "SoupXfull")
  seu <- add_assay(seu, soupx_reduced,   "SoupXreduced")
  seu <- add_assay(seu, decontx_full,    "DecontXfull")
  seu <- add_assay(seu, decontx_reduced, "DecontXreduced")

  return(seu)
}
```

## Python-based methods

```{r}
#' Read a corrected count matrix from an H5 or H5AD file
#'
#' Parses sparse matrix indices via rhdf5. Supports CellBender (H5),
#' scAR (H5AD), and CellClear (H5AD) output formats.
#'
#' @param path Character. Path to the H5 or H5AD file.
#' @param type Character. One of "CellBender", "scAR", or "CellClear"
#'   (default "CellBender").
#' @param name Character. Sample name appended to barcodes for uniqueness.
#'
#' @return A sparse dgCMatrix (genes x barcodes).
read_H5AD <- function(path, type = "CellBender", name) {
  root <- rhdf5::H5Fopen(path)

  if (type %in% c("scAR", "CellClear")) {
    i <- as.vector(unlist(rhdf5::h5read(root, "/X/indices")))
    p <- as.vector(unlist(rhdf5::h5read(root, "/X/indptr")))
    x <- as.vector(unlist(rhdf5::h5read(root, "/X/data")))
    o <- as.vector(rhdf5::h5read(root, "/obs")$"_index")
    v <- as.vector(rhdf5::h5read(root, "/var")$"_index")
  } else if (type == "CellBender") {
    i <- as.vector(unlist(rhdf5::h5read(root, "/matrix/indices")))
    p <- as.vector(unlist(rhdf5::h5read(root, "/matrix/indptr")))
    x <- as.vector(unlist(rhdf5::h5read(root, "/matrix/data")))
    o <- as.vector(rhdf5::h5read(root, "/matrix/barcodes"))
    v <- as.vector(rhdf5::h5read(root, "/matrix/features/name"))
  }

  rhdf5::H5Fclose(root)

  dims   <- c(length(v), length(o))
  counts <- Matrix::sparseMatrix(i = i, p = p, x = x, index1 = FALSE, dims = dims)
  rownames(counts) <- v
  colnames(counts) <- paste(o, name, sep = "_")
  return(counts)
}
```

```{r}
#' Add a list of corrected count matrices as a new Seurat assay
#'
#' Aligns each matrix to the features and barcodes in `object`, padding
#' missing genes with zeros, then inserts the combined matrix as a new
#' assay and adds species fraction metadata columns.
#'
#' @param object A Seurat object.
#' @param mat Named list of corrected count matrices (one per sample).
#' @param name Character. Name for the new assay (e.g. "CellBender").
#'
#' @return The Seurat object with the new assay added.
add_assay <- function(object, mat, name) {

  for (dataset in seq_along(mat)) {
    rownames(mat[[dataset]]) <- gsub("_", "-", rownames(mat[[dataset]]))
    mat[[dataset]] <- mat[[dataset]][
      rownames(mat[[dataset]]) %in% rownames(object@assays$RNA$counts), 
    ]
    mat[[dataset]] <- mat[[dataset]][!duplicated(rownames(mat[[dataset]])), ]

    # Pad with zero rows for genes absent from this sample's output
    if (nrow(object) > nrow(mat[[dataset]])) {
      missing_genes  <- rownames(object)[
        !(rownames(object) %in% rownames(mat[[dataset]]))
      ]
      missing_counts <- Matrix::Matrix(
        0,
        nrow     = length(missing_genes),
        ncol     = ncol(mat[[dataset]]),
        sparse   = TRUE,
        dimnames = list(missing_genes, colnames(mat[[dataset]]))
      )
      mat[[dataset]] <- rbind(missing_counts, mat[[dataset]])
    }

    mat[[dataset]] <- mat[[dataset]][
      match(rownames(object@assays$RNA$counts), rownames(mat[[dataset]])), 
    ]
  }

  mat <- do.call("cbind", mat)

  # Restrict to barcodes present in both object and corrected matrix
  barcodes.keep <- intersect(object$barcode, colnames(mat))
  object        <- subset(object, cells = barcodes.keep)
  mat           <- mat[, match(object$barcode, colnames(mat))]
  colnames(mat) <- colnames(object)

  # Insert assay and add species fraction columns
  object[[name]] <- CreateAssayObject(mat, min.cells = 0, min.features = 0)
  object[[paste0(name, ".percent.mm10")]]   <- PercentageFeatureSet(
    object, pattern = "^mm10",   assay = name
  )[, 1]
  object[[paste0(name, ".percent.GRCh38")]] <- PercentageFeatureSet(
    object, pattern = "^GRCh38", assay = name
  )[, 1]

  return(object)
}
```

------------------------------------------------------------------------

# Analysing results from ambient RNA removal

```{r}
#' Calculate per-cell and per-gene exon fractions
#'
#' Compares Gene (exon-only) and GeneFull_Ex50pAS (full gene body) STARsolo
#' output folders to derive the fraction of UMIs from exonic regions.
#'
#' @param path Character. Path to the STARsolo Solo.out directory (must
#'   contain GeneFull_Ex50pAS/raw/ and Gene/raw/ subdirectories).
#' @param object A Seurat object used to match barcodes.
#' @param name Character. Sample name used in barcode construction.
#' @param paste1 Logical. If TRUE (default) barcodes are formatted as
#'   "<barcode>-1_<name>"; if FALSE, "<barcode>_<name>".
#'
#' @return A named list with elements:
#'   \describe{
#'     \item{cells}{Per-cell exon fractions (named numeric vector).}
#'     \item{features}{Per-gene exon fractions (named numeric vector).}
#'   }
calculate_exon_fraction <- function(path, object, name, paste1 = TRUE) {
  total <- load_folder(file.path(path, "GeneFull_Ex50pAS", "raw"))
  exon  <- load_folder(file.path(path, "Gene", "raw"))

  total <- total[rownames(total) %in% rownames(exon), ]
  exon  <- exon[rownames(total), ]

  if (paste1) {
    colnames(total) <- paste0(colnames(total), "-1_", name)
    colnames(exon)  <- paste0(colnames(exon),  "-1_", name)
  } else {
    colnames(total) <- paste0(colnames(total), "_", name)
    colnames(exon)  <- paste0(colnames(exon),  "_", name)
  }

  total <- total[, colnames(total) %in% colnames(object)]
  exon  <- exon[,  colnames(exon)  %in% colnames(object)]

  list(
    cells    = colSums(exon) / colSums(total),
    features = rowSums(exon) / rowSums(total)
  )
}
```

```{r}
#' Estimate ambient background expression from empty droplets
#'
#' Identifies droplets with total UMI count <= `threshold`, sums their
#' counts per gene, and returns the result aligned to `object`.
#'
#' @param path Character. Path to a count matrix folder readable by
#'   \code{load_folder()}.
#' @param object A Seurat object used to match the gene set.
#' @param threshold Integer. Max UMI count to treat a droplet as empty
#'   (default 50).
#'
#' @return Named numeric vector of summed ambient UMI counts per gene,
#'   ordered to match \code{rownames(object)}.
calculate_background_profile <- function(path, object, threshold = 50) {
  counts <- load_folder(path)

  background_barcodes <- names(which(colSums(counts) <= threshold))
  background          <- counts[, colnames(counts) %in% background_barcodes]

  rownames(background) <- gsub("_",   "-", rownames(background))
  rownames(background) <- gsub("---", "-", rownames(background))
  background <- background[rownames(background) %in% rownames(object), ]
  background <- background[!duplicated(rownames(background)), ]

  # Pad missing genes with zeros
  missing_genes  <- rownames(object)[!(rownames(object) %in% rownames(background))]
  missing_counts <- Matrix::Matrix(
    0, nrow = length(missing_genes), ncol = ncol(background), sparse = TRUE,
    dimnames = list(missing_genes, colnames(background))
  )
  background <- rbind(missing_counts, background)
  background <- background[match(rownames(object), rownames(background)), ]

  rowSums(background)
}
```

```{r}
#' Calculate per-cell ambient RNA load percentages
#'
#' Computes the ambient fraction (as a percentage) per cell. When `samples`
#' and `seu` are supplied, results are grouped by sample and returned as a
#' data frame suitable for ridge plots.
#'
#' @param ambient_list Named list of ambient count matrices.
#' @param raw_list Named list of total (raw) count matrices. Must have the
#'   same names as `ambient_list`.
#' @param samples Optional character vector of sample names. When provided,
#'   results are grouped by sample.
#' @param seu Optional Seurat object whose metadata is used to assign cells
#'   to samples (required when `samples` is not NULL).
#'
#' @return If `samples` is NULL: a numeric vector of per-cell ambient
#'   percentages. Otherwise: a data frame with columns `group` (sample
#'   label) and `value` (ambient percentage per cell).
calc_ambient_load_percentages <- function(ambient_list, raw_list,
                                           samples = NULL, seu = NULL) {
  if (is.null(names(ambient_list)) || any(names(ambient_list) == "")) {
    stop("All elements of ambient_list must be named.")
  }
  if (is.null(names(raw_list)) || any(names(raw_list) == "")) {
    stop("All elements of raw_list must be named.")
  }
  if (!setequal(names(ambient_list), names(raw_list))) {
    stop("ambient_list and raw_list must have the same names.")
  }

  raw_list <- raw_list[names(ambient_list)]

  if (is.null(samples)) {
    ambient_load <- numeric()
    for (name in names(ambient_list)) {
      ambient_load <- c(
        ambient_load,
        (colSums(ambient_list[[name]], na.rm = TRUE) /
           colSums(raw_list[[name]],   na.rm = TRUE)) * 100
      )
    }
    return(ambient_load[!is.na(ambient_load)])

  } else {
    ambient_load_list <- list()
    for (i in seq_along(samples)) {
      sample_cols <- rownames(seu@meta.data)[seu$orig.ident == samples[i]]
      values      <- numeric()
      for (name in names(ambient_list)) {
        values <- c(
          values,
          (colSums(ambient_list[[name]][, colnames(ambient_list[[name]]) %in% sample_cols, drop = FALSE], na.rm = TRUE) /
             colSums(raw_list[[name]][,   colnames(raw_list[[name]])    %in% sample_cols, drop = FALSE], na.rm = TRUE)) * 100
        )
      }
      ambient_load_list[[paste0("S", i)]] <- values[!is.na(values)]
    }

    df <- data.frame(
      group = rep(names(ambient_load_list), lengths(ambient_load_list)),
      value = unlist(ambient_load_list, use.names = FALSE)
    )
    return(df[!is.na(df$value), ])
  }
}
```

```{r}
#' Match ambient load with exon fractions (medium complexity dataset)
#'
#' Builds a combined data frame of per-gene exon fractions and log-ambient
#' expression for use in scatter plots, handling the cross-species design
#' of the medium complexity dataset separately for human and mouse samples.
#'
#' @param exon_fraction_per_feature_1 Named numeric vector of per-gene exon
#'   fractions for the first sample group (human samples).
#' @param exon_fraction_per_feature_2 Named numeric vector of per-gene exon
#'   fractions for the second sample group (mouse samples).
#' @param raw_human_ambient Sparse matrix of ambient counts from human
#'   samples (GRCh38 genes contaminating mouse spike-in cells).
#' @param raw_mouse_ambient Sparse matrix of ambient counts from mouse
#'   samples (mm10 genes contaminating human spike-in cells).
#'
#' @return A data frame with columns `exon_fraction`, `log_ambient_expression`,
#'   and `species`.
match_ambient_load_with_exon_fractions_mc <- function(exon_fraction_per_feature_1,
                                                       exon_fraction_per_feature_2,
                                                       raw_human_ambient,
                                                       raw_mouse_ambient) {
  # Human samples: ambient signal is from GRCh38 genes
  ef_human <- exon_fraction_per_feature_1[grepl("^GRCh38", names(exon_fraction_per_feature_1))]
  ef_human <- ef_human[names(ef_human) %in% rownames(raw_human_ambient)]
  raw_human_ambient <- raw_human_ambient[
    rownames(raw_human_ambient) %in% names(exon_fraction_per_feature_1), 
  ]
  raw_human_ambient <- raw_human_ambient[match(names(ef_human), rownames(raw_human_ambient)), ]

  df_human <- data.frame(
    exon_fraction          = ef_human,
    log_ambient_expression = log1p(rowSums(raw_human_ambient)),
    species                = "human"
  )

  # Mouse samples: ambient signal is from mm10 genes
  ef_mouse <- exon_fraction_per_feature_2[grepl("^mm10", names(exon_fraction_per_feature_2))]
  ef_mouse <- ef_mouse[names(ef_mouse) %in% rownames(raw_mouse_ambient)]
  raw_mouse_ambient <- raw_mouse_ambient[
    rownames(raw_mouse_ambient) %in% names(exon_fraction_per_feature_2), 
  ]
  raw_mouse_ambient <- raw_mouse_ambient[match(names(ef_mouse), rownames(raw_mouse_ambient)), ]

  df_mouse <- data.frame(
    exon_fraction          = ef_mouse,
    log_ambient_expression = log1p(rowSums(raw_mouse_ambient)),
    species                = "mouse"
  )

  rbind(df_human, df_mouse)
}
```

```{r}
#' Estimate per-cell, per-gene ambient RNA from genotype SNP data
#'
#' For a single genotype sample, reads the intermediate SNP calling results
#' (csc_intermediate.RDS), filters to high-confidence CAST-specific SNPs
#' (>= 2 supporting reads), and calculates the ambient fraction per unique
#' gene-in-cell combination as ALT / total reads.
#'
#' Used only for the genotype (strain-mixing) dataset.
#'
#' @param sample Character. Sample name matching the directory structure
#'   under data/processed/genotype/genotype_estimation/.
#'
#' @return A data frame with columns Barcode, Symbol, Ambient, and Position.
calc_ambient_from_SNPs <- function(sample) {
  sample_annotation <- readRDS(here::here(paste0(
    "data/processed/genotype/genotype_estimation/", sample, "/csc_intermediate.RDS"
  )))

  # Retain only CAST (Castaneus) strain SNPs
  sample_annotation <- sample_annotation[sample_annotation$Strain == "CAST", ]

  # Require >= 2 reads for confidence
  sample_annotation <- sample_annotation[sample_annotation$DP >= 2, ]

  # Keep only rows where reference and alternative allele differ (= ambient)
  sample_annotation <- sample_annotation[
    sample_annotation$Strain != sample_annotation$SNPisALT, 
  ]

  # Match barcode format to Seurat object
  sample_annotation$cell <- paste(sample_annotation$cell, sample, sep = "_")
  sample_annotation <- sample_annotation[
    sample_annotation$cell %in% gen_seu$barcode, 
  ]

  # Remove SNPs assigned to more than one species in SNPisALT
  keep <- vapply(seq_len(nrow(sample_annotation)), function(i) {
    !grepl(sample_annotation$Strain[i], sample_annotation$SNPisALT[i])
  }, logical(1))
  sample_annotation <- sample_annotation[keep, ]

  # Calculate ambient fraction per unique gene-in-cell combination
  sample_annotation$ID <- paste(sample_annotation$cell, sample_annotation$SYMBOL, sep = ":")

  res <- do.call(rbind, lapply(unique(sample_annotation$ID), function(id) {
    rows  <- sample_annotation[sample_annotation$ID == id, ]
    cell  <- sub(":.*", "", id)
    gene  <- sub(".*:", "", id)
    ALT   <- sum(rows$AD)
    TOTAL <- sum(rows$DP)
    data.frame(
      Barcode  = cell,
      Symbol   = gene,
      Ambient  = ALT / TOTAL,
      Position = rows$position[1]
    )
  }))

  return(res)
}
```

```{r}
#' Run ambient or endogenous removal analysis across all tools and species
#'
#' For a given dataset and compartment, extracts per-tool corrected counts
#' for human and mouse cells, computes the fraction of counts removed, saves
#' per-tool scatter plots, and writes a summary table.
#'
#' @param seu A Seurat object containing all decontamination assays.
#' @param dataset_name Character. Label used in output filenames.
#' @param compartment Character. Either "ambient" or "endogenous".
#' @param plot_dir Character. Directory for output PDF plots.
#' @param table_path Character. File path for the output .xlsx table.
#'
#' @return A named list with elements:
#'   \describe{
#'     \item{plots}{Named list of ggplot objects.}
#'     \item{data}{Combined data frame of all tools and species.}
#'     \item{per_tool_species_dfs}{Per-tool, per-species data frames.}
#'   }
run_compartment_removal_analysis <- function(seu,
                                              dataset_name,
                                              compartment = c("ambient", "endogenous"),
                                              plot_dir,
                                              table_path) {
  compartment <- match.arg(compartment)

  tools      <- c("CellBender", "CellClear", "FastCAR", "scAR", "scCDC",
                  "SoupX", "SoupX_r", "DecontX", "DecontX_r")
  tools_abbr <- c("cb", "cc", "fc", "ar", "cd", "sx", "sxr", "dx", "dxr")
  tools_map  <- setNames(tools_abbr, tools)
  input_name <- paste0("input_", compartment)

  # Extract corrected counts for each species
  corrected <- list(
    human = divide_ambient_and_endogenous_per_tool(seu = seu, extract = "human"),
    mouse = divide_ambient_and_endogenous_per_tool(seu = seu, extract = "mouse")
  )

  # Build per-tool, per-species data frames of fraction removed
  compartment_dfs <- list()
  for (species in names(corrected)) {
    for (tool in tools) {
      abbr      <- tools_map[[tool]]
      tool_name <- paste0(abbr, "_", compartment)
      compartment_dfs[[paste(tool, species, sep = "_")]] <- data.frame(
        log_UMIs         = log1p(rowSums(corrected[[species]][[input_name]])),
        fraction_removed = (rowSums(corrected[[species]][[input_name]]) -
                              rowSums(corrected[[species]][[tool_name]])) /
          rowSums(corrected[[species]][[input_name]]),
        species = species
      )
    }
  }

  get_plot_df <- function(tool) {
    rbind(
      compartment_dfs[[paste(tool, "human", sep = "_")]],
      compartment_dfs[[paste(tool, "mouse", sep = "_")]]
    )
  }

  # Generate scatter plots
  plots <- setNames(lapply(seq_along(tools), function(i) {
    ggplot(get_plot_df(tools[[i]]), aes(x = log_UMIs, y = fraction_removed)) +
      geom_point_rast(
        colour    = custom_colors[[i]],
        size      = 0.5,
        stroke    = 0,
        alpha     = 0.5,
        raster.dpi = 1500
      ) +
      plot_theme() +
      labs(x = NULL, y = NULL) +
      ylim(-1, 1)
  }), paste0(dataset_name, "_", compartment, "_removal_", tools))

  # Save plots
  dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
  for (plot_name in names(plots)) {
    ggsave(
      filename = file.path(plot_dir, paste0(plot_name, ".pdf")),
      plot     = plots[[plot_name]],
      dpi      = 300,
      width    = 35,
      height   = 35,
      units    = "mm"
    )
  }

  # Combine into a single table
  final_df <- do.call(rbind, lapply(tools, function(tool) {
    df <- get_plot_df(tool)
    df$decontamination_method <- tool
    df
  }))
  final_df$dataset    <- dataset_name
  final_df$compartment <- compartment
  write_xlsx(final_df, path = table_path)

  list(plots = plots, data = final_df, per_tool_species_dfs = compartment_dfs)
}
```

```{r}
#' Evaluate sample integration quality across decontamination methods
#'
#' Used only for the medium and high complexity datasets. For each assay in
#' `object_list`, runs the Seurat pipeline (optionally with Harmony) and
#' computes LISI scores across a range of PCA dimensions.
#'
#' @param object_list Named list of Seurat objects, one per method.
#' @param assay Unused placeholder (kept for API compatibility).
#' @param integrate Logical. Use Harmony integration (default TRUE).
#' @param min.dim Integer. Minimum dimensions to evaluate (default 5).
#' @param max.dim Integer. Maximum dimensions to evaluate (default 20).
#' @param labels Character. Metadata column for cell type labels
#'   (default "labels").
#' @param batches Character. Metadata column for batch labels
#'   (default "orig.ident").
#'
#' @return A named list with element `lisi`: a data frame of LISI scores
#'   across methods and dimensions.
evaluate_integration <- function(object_list,
                                  assay,
                                  integrate = TRUE,
                                  min.dim   = 5,
                                  max.dim   = 20,
                                  labels    = "labels",
                                  batches   = "orig.ident") {
  lisi_res     <- as.data.frame(matrix(ncol = 5, nrow = 0))
  lisi_counter <- 1

  for (dataset in seq_along(object_list)) {
    DefaultAssay(object_list[[dataset]]) <- names(object_list)[dataset]

    if (names(object_list)[dataset] != "RNA") {
      object_list[[dataset]][["RNA"]] <- object_list[[dataset]][[names(object_list)[dataset]]]
    }
    DefaultAssay(object_list[[dataset]]) <- "RNA"

    new_seu <- NormalizeData(object_list[[dataset]], verbose = FALSE)

    if (integrate) {
      new_seu_list <- SplitObject(new_seu, split.by = "orig.ident")
      VariableFeatures(new_seu) <- SelectIntegrationFeatures(new_seu_list, verbose = FALSE)
    } else {
      new_seu <- FindVariableFeatures(new_seu, verbose = FALSE)
    }

    new_seu <- ScaleData(new_seu, verbose = FALSE)
    new_seu <- RunPCA(new_seu, verbose = FALSE)

    if (integrate) {
      new_seu <- RunHarmony(new_seu, group.by.vars = "orig.ident", verbose = FALSE)
    }

    nlabel <- length(unique(new_seu@meta.data[, labels]))
    nbatch <- length(unique(new_seu@meta.data[, batches]))

    for (i in min.dim:max.dim) {
      embeddings <- if (integrate) {
        new_seu@reductions$harmony@cell.embeddings[, 1:i]
      } else {
        new_seu@reductions$pca@cell.embeddings[, 1:i]
      }

      lisi <- compute_lisi(embeddings, new_seu@meta.data,
                           label_colnames = c("orig.ident", "labels"))
      lisi[, 1] <- (lisi[, 1] - 1) / (nbatch - 1)
      lisi[, 2] <- (nlabel - lisi[, 2]) / (nlabel - 1)

      lisi_res[lisi_counter, ] <- list(
        names(object_list)[dataset],
        i,
        ifelse(integrate, "harmony", "pca"),
        mean(lisi[, 1]),
        mean(lisi[, 2])
      )
      lisi_counter <- lisi_counter + 1
    }
  }

  list(lisi = lisi_res)
}
```

------------------------------------------------------------------------

# Figure helpers

## Plot theme

```{r}
#' Base ggplot2 theme for publication figures
#'
#' @param fixed_aspect Logical. If TRUE (default), enforces aspect.ratio = 1.
#'
#' @return A ggplot2 theme object.
plot_theme <- function(fixed_aspect = TRUE) {
  base_theme <- theme(
    panel.background  = element_rect(fill = "white", colour = NA),
    plot.background   = element_rect(fill = "white", colour = NA),
    plot.title        = element_blank(),
    axis.text.y       = element_text(margin = margin(r = 0.5, unit = "mm")),
    axis.text.x       = element_text(margin = margin(t = 0.6, unit = "mm")),
    axis.title.x      = element_text(margin = margin(t = 0.3, unit = "mm")),
    axis.title.y      = element_text(margin = margin(r = 0.6, unit = "mm")),
    axis.title        = element_text(size = 7),
    axis.text         = element_text(size = 5, colour = "black"),
    axis.line         = element_line(linewidth = 0.4),
    axis.ticks        = element_line(lineend = "square", linewidth = 0.4, color = "black"),
    axis.ticks.length = unit(0.5, "mm"),
    legend.key.size   = unit(0.3, "cm"),
    legend.title      = element_text(size = 5, margin = margin(l = 1)),
    legend.text       = element_text(size = 5, margin = margin(l = 1)),
    legend.position   = "right",
    legend.box.margin = margin(-15, -15, -15, -15),
    legend.margin     = margin(0, 0, 0, 0),
    legend.justification = "left",
    legend.spacing.x  = unit(0, "mm"),
    legend.key.width  = unit(1, "mm"),
    text              = element_text(colour = "black")
  )

  if (fixed_aspect) base_theme <- base_theme + theme(aspect.ratio = 1)
  base_theme
}
```

## Annotation helper

```{r}
#' Add text annotation at publication font size
#'
#' Thin wrapper around \code{ggplot2::annotate()} that sets the default
#' size to 5/.pt (approximately 5 pt).
#'
#' @param ... Arguments passed to \code{annotate()}.
#' @param size Numeric. Text size in ggplot units (default 5/.pt).
general_annotation <- function(..., size = 5 / .pt) {
  annotate(..., size = size)
}
```

## Remove point stroke

```{r}
#' Remove the stroke from the first layer of a ggplot
#'
#' Useful for reducing visual weight of small rasterised points.
#'
#' @param figure A ggplot object.
#'
#' @return The modified ggplot object.
remove_stroke <- function(figure) {
  figure$layers[[1]]$aes_params$stroke <- 0
  figure
}
```

## Axis labels

```{r}
# Standard UMAP axis labels
UMAP_labs <- labs(x = "UMAP 1", y = "UMAP 2")
```

## Legend guides

```{r}
#' Guide settings for a continuous colour legend
#'
#' @param legend_title Character. Legend title (default "Title").
#' @param barwidth Unit. Width of the colour bar (default 0.2 lines).
#' @param barheight Numeric. Height of the colour bar (default 4).
guide_continuous <- function(legend_title = "Title",
                              barwidth     = unit(0.2, "lines"),
                              barheight    = 4) {
  guides(
    color = guide_colorbar(
      title          = legend_title,
      title.position = "left",
      barwidth       = barwidth,
      barheight      = barheight,
      direction      = "vertical",
      title.hjust    = 0.5,
      label.theme    = element_text(margin = margin(t = 0, r = 0, b = 0, l = 1)),
      title.theme    = element_text(angle = 90, vjust = 0.5)
    )
  )
}
```

```{r}
#' Guide settings for a discrete colour / fill legend
#'
#' @param ncol Integer. Number of legend columns (default 1).
discrete_guide_settings <- function(ncol = 1) {
  guide_legend(
    direction      = "vertical",
    label.theme    = element_text(margin = margin(r = -3, l = 1), vjust = 0.5),
    override.aes   = list(size = 1),
    keyheight      = unit(0.5, "lines"),
    keywidth       = unit(0.2, "lines"),
    label.position = "right",
    label.hjust    = 0,
    label.vjust    = 0.5,
    ncol           = ncol
  )
}

guide_discrete <- function(ncol = 1) {
  guides(
    color = discrete_guide_settings(ncol),
    fill  = discrete_guide_settings(ncol)
  )
}
```

```{r}
#' Combined continuous colour + discrete size guide
#'
#' @param legend_title Character. Title for the colour bar (default "Title").
#' @param barwidth Unit. Colour bar width (default 0.2 lines).
#' @param barheight Numeric. Colour bar height (default 4).
#' @param size_ncol Integer. Columns for the size legend (default 1).
guide_combined <- function(legend_title = "Title",
                            barwidth     = unit(0.2, "lines"),
                            barheight    = 4,
                            size_ncol    = 1) {
  guides(
    color = guide_colorbar(
      title          = legend_title,
      title.position = "left",
      barwidth       = barwidth,
      barheight      = barheight,
      direction      = "vertical",
      title.hjust    = 0.5,
      label          = FALSE,
      label.theme    = element_text(margin = margin(t = 0, r = 0, b = 0, l = 1)),
      title.theme    = element_text(angle = 90, vjust = 0.5)
    ),
    size = guide_legend(
      direction      = "vertical",
      label.theme    = element_text(margin = margin(r = -3, l = 1), vjust = 0.5),
      override.aes   = list(size = 1),
      label.position = "right",
      label.hjust    = 0,
      label.vjust    = 0.5,
      ncol           = size_ncol
    )
  )
}
```

## Colour palette

```{r}
# Colourblind-friendly palette tested across all figures
custom_colors <- c(
  "#0D0887",  # deep purple
  "#ED7953",  # orange
  "#FDB32F",  # yellow
  "#81D8D0",  # teal
  "#C8A2C8",  # lilac
  "#66C2A5",  # green
  "#8DA0CB",  # blue-grey
  "#9C179E",  # magenta
  "#E78AC3",  # pink
  "#A6D854",  # lime
  "#FFD92F",  # bright yellow
  "#5E4FA2",  # indigo
  "#3288BD",  # blue
  "#F46D43",  # coral
  "#FDAE61",  # light orange
  "#1A9850"   # dark green
)
```
