---
title: "04_preprocessing_and_quality_control"
format: pdf
editor: visual
---

All count matrices are filtered for cell-containing droplets and low-quality cells. emptyDrops (FDR ≤ 0.05) is used for all droplet-based datasets; Smart-seq2 is filtered manually using fixed count and feature thresholds (emptyDrops is not applicable to plate-based data). CellBender performs its own filtering internally, so only the raw matrices are passed to it.

------------------------------------------------------------------------

# Create output directory structure

```{r}
main_dir <- here::here()

dir.create(file.path(main_dir, "data/processed/low_complexity/emptyDrops_filtered/rds-files"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(main_dir, "data/processed/low_complexity/emptyDrops_filtered/h5-files"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(main_dir, "data/processed/medium_complexity/emptyDrops_filtered/rds-files"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(main_dir, "data/processed/medium_complexity/emptyDrops_filtered/h5-files"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(main_dir, "data/processed/high_complexity/emptyDrops_filtered/rds-files"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(main_dir, "data/processed/high_complexity/emptyDrops_filtered/h5-files"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(main_dir, "data/processed/genotype/emptyDrops_filtered/rds-files"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(main_dir, "data/processed/genotype/emptyDrops_filtered/h5-files"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(main_dir, "data/processed/smart-seq2/filtered/rds-files"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(main_dir, "data/processed/smart-seq2/filtered/h5-files"),        recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(main_dir, "data/processed/synthetic/filtered/rds-files"),        recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(main_dir, "data/processed/synthetic/filtered/h5-files"),         recursive = TRUE, showWarnings = FALSE)
```

------------------------------------------------------------------------

# Load count matrices

```{r}
lc_count_matrix <- readRDS(here::here("data/raw/low_complexity/count_matrix/lc_count_matrix.Rds"))
mc_human_count_matrices <- readRDS(here::here("data/raw/medium_complexity/count_matrices/mc_human_count_matrices.Rds"))
mc_mouse_count_matrices <- readRDS(here::here("data/raw/medium_complexity/count_matrices/mc_mouse_count_matrices.Rds"))
hc_count_matrices <- readRDS(here::here("data/raw/high_complexity/count_matrices/hc_count_matrices.Rds"))
gen_count_matrices <- readRDS(here::here("data/raw/genotype/count_matrices/gen_count_matrices.Rds"))
smart_count_matrices <- readRDS(here::here("data/raw/smart-seq2/count_matrices/smart_count_matrices.Rds"))
```

------------------------------------------------------------------------

# Quality control

## Low complexity

emptyDrops lower UMI threshold set to 500 (single-sample dataset).

```{r}
set.seed(100)

lc_ed <- as.data.frame(DropletUtils::emptyDrops(m = lc_count_matrix, lower = 500))
lc_ed <- lc_ed[!is.na(lc_ed$FDR), ]

# Cell barcodes passing FDR <= 0.05
lc_barcodes_post_ed <- rownames(lc_ed[lc_ed$FDR <= 0.05, ])

saveRDS(lc_barcodes_post_ed, file = here::here("data/processed/low_complexity/emptyDrops_filtered/rds-files/low_complexity_barcodes_post_emptyDrops.Rds"))

# Filtered matrix for Python-based methods
lc_filtered <- lc_count_matrix[, colnames(lc_count_matrix) %in% lc_barcodes_post_ed]
lc_filtered <- lc_filtered[rowSums(lc_filtered) > 0, ]

write10xCounts(
  here::here("data/processed/low_complexity/emptyDrops_filtered/h5-files/low_complexity_post_emptyDrops.h5"),
  lc_filtered,
  barcodes = colnames(lc_filtered),
  gene.id = rownames(lc_filtered),
  gene.symbol = rownames(lc_filtered),
  gene.type = "Gene Expression",
  overwrite = TRUE,
  type = "HDF5",
  genome = "unknown",
  version = "3",
  chemistry = "Single Cell 3' v2",
  original.gem.groups = 1L,
  library.ids = "custom"
)
```

## Medium complexity

emptyDrops lower UMI threshold set to 200 for all medium complexity samples.

### Human samples

```{r}
mc_human_barcodes_post_ed <- list()
mc_human_filtered_matrices <- list()

for (i in seq_along(mc_human_count_matrices)) {
  set.seed(100)

  ed <- as.data.frame(DropletUtils::emptyDrops(
    m = mc_human_count_matrices[[i]], lower = 200
  ))
  ed <- ed[!is.na(ed$FDR), ]

  mc_human_barcodes_post_ed[[i]] <- rownames(ed[ed$FDR <= 0.05, ])
  names(mc_human_barcodes_post_ed)[i] <- mc_human_samples[i]

  saveRDS(mc_human_barcodes_post_ed[[i]], here::here(paste0(
    "data/processed/medium_complexity/emptyDrops_filtered/rds-files/",
    mc_human_samples[i], "_barcodes_post_emptyDrops.Rds"
  )))

  # Filtered matrix for Python-based methods
  mat <- mc_human_count_matrices[[i]]
  mat <- mat[, colnames(mat) %in% mc_human_barcodes_post_ed[[i]]]
  mat <- mat[rowSums(mat) > 0, ]
  mc_human_filtered_matrices[[i]] <- mat

  write10xCounts(
    here::here(paste0(
      "data/processed/medium_complexity/emptyDrops_filtered/h5-files/",
      mc_human_samples[i], "_post_emptyDrops.h5"
    )),
    mat,
    barcodes = colnames(mat),
    gene.id = rownames(mat),
    gene.symbol = rownames(mat),
    gene.type = "Gene Expression",
    overwrite = TRUE,
    type = "HDF5",
    genome = "unknown",
    version = "3",
    chemistry = "Single Cell 3' v2",
    original.gem.groups = 1L,
    library.ids = "custom"
  )

  message("Done: ", mc_human_samples[i])
}

saveRDS(mc_human_barcodes_post_ed, file = here::here("data/processed/medium_complexity/emptyDrops_filtered/rds-files/medium_complexity_human_barcodes_post_ed.Rds"))
```

### Mouse samples

```{r}
mc_mouse_barcodes_post_ed <- list()
mc_mouse_filtered_matrices <- list()

for (i in seq_along(mc_mouse_count_matrices)) {
  set.seed(100)

  ed <- as.data.frame(DropletUtils::emptyDrops(
    m = mc_mouse_count_matrices[[i]], lower = 200
  ))
  ed <- ed[!is.na(ed$FDR), ]

  mc_mouse_barcodes_post_ed[[i]] <- rownames(ed[ed$FDR <= 0.05, ])
  names(mc_mouse_barcodes_post_ed)[i] <- mc_mouse_samples[i]

  saveRDS(mc_mouse_barcodes_post_ed[[i]], here::here(paste0(
    "data/processed/medium_complexity/emptyDrops_filtered/rds-files/",
    mc_mouse_samples[i], "_barcodes_post_emptyDrops.Rds"
  )))

  mat <- mc_mouse_count_matrices[[i]]
  mat <- mat[, colnames(mat) %in% mc_mouse_barcodes_post_ed[[i]]]
  mat <- mat[rowSums(mat) > 0, ]
  mc_mouse_filtered_matrices[[i]] <- mat

  write10xCounts(
    here::here(paste0(
      "data/processed/medium_complexity/emptyDrops_filtered/h5-files/",
      mc_mouse_samples[i], "_post_emptyDrops.h5"
    )),
    mat,
    barcodes = colnames(mat),
    gene.id = rownames(mat),
    gene.symbol = rownames(mat),
    gene.type = "Gene Expression",
    overwrite = TRUE,
    type = "HDF5",
    genome = "unknown",
    version = "3",
    chemistry = "Single Cell 3' v2",
    original.gem.groups = 1L,
    library.ids = "custom"
  )

  message("Done: ", mc_mouse_samples[i])
}

saveRDS(mc_mouse_barcodes_post_ed, file = here::here("data/processed/medium_complexity/emptyDrops_filtered/rds-files/medium_complexity_mouse_barcodes_post_ed.Rds"))
```

## High complexity

emptyDrops lower UMI threshold set to 200.

```{r}
hc_barcodes_post_ed <- list()
hc_filtered_matrices <- list()

for (i in seq_along(hc_count_matrices)) {
  set.seed(100)

  ed <- as.data.frame(DropletUtils::emptyDrops(
    m = hc_count_matrices[[i]], lower = 200
  ))
  ed <- ed[!is.na(ed$FDR), ]

  hc_barcodes_post_ed[[i]] <- rownames(ed[ed$FDR <= 0.05, ])
  names(hc_barcodes_post_ed)[i] <- hc_samples[i]

  saveRDS(hc_barcodes_post_ed[[i]], here::here(paste0(
    "data/processed/high_complexity/emptyDrops_filtered/rds-files/",
    hc_samples[i], "_barcodes_post_emptyDrops.Rds"
  )))

  mat <- hc_count_matrices[[i]]
  mat <- mat[, colnames(mat) %in% hc_barcodes_post_ed[[i]]]
  mat <- mat[rowSums(mat) > 0, ]
  hc_filtered_matrices[[i]] <- mat

  write10xCounts(
    here::here(paste0(
      "data/processed/high_complexity/emptyDrops_filtered/h5-files/",
      hc_samples[i], "_post_emptyDrops.h5"
    )),
    mat,
    barcodes = colnames(mat),
    gene.id = rownames(mat),
    gene.symbol = rownames(mat),
    gene.type = "Gene Expression",
    overwrite = TRUE,
    type = "HDF5",
    genome = "unknown",
    version = "3",
    chemistry = "Single Cell 3' v2",
    original.gem.groups = 1L,
    library.ids = "custom"
  )

  message("Done: ", hc_samples[i])
}

saveRDS(hc_barcodes_post_ed, file = here::here("data/processed/high_complexity/emptyDrops_filtered/rds-files/high_complexity_barcodes_post_ed.Rds"))
```

## Genotype

emptyDrops lower UMI threshold set to 700 (nuclei dataset, lower library complexity than whole-cell datasets).

```{r}
gen_barcodes_post_ed  <- list()
gen_filtered_matrices <- list()

for (i in seq_along(gen_count_matrices)) {
  set.seed(100)

  ed <- as.data.frame(DropletUtils::emptyDrops(
    m = gen_count_matrices[[i]], lower = 700
  ))
  ed <- ed[!is.na(ed$FDR), ]

  gen_barcodes_post_ed[[i]] <- rownames(ed[ed$FDR <= 0.05, ])
  names(gen_barcodes_post_ed)[i] <- gen_samples[i]

  saveRDS(gen_barcodes_post_ed[[i]], here::here(paste0(
    "data/processed/genotype/emptyDrops_filtered/rds-files/",
    gen_samples[i], "_barcodes_post_emptyDrops.Rds"
  )))

  mat <- gen_count_matrices[[i]]
  mat <- mat[, colnames(mat) %in% gen_barcodes_post_ed[[i]]]
  mat <- mat[rowSums(mat) > 0, ]
  gen_filtered_matrices[[i]] <- mat

  write10xCounts(
    here::here(paste0(
      "data/processed/genotype/emptyDrops_filtered/h5-files/",
      gen_samples[i], "_post_emptyDrops.h5"
    )),
    mat,
    barcodes = colnames(mat),
    gene.id = rownames(mat),
    gene.symbol = rownames(mat),
    gene.type = "Gene Expression",
    overwrite = TRUE,
    type = "HDF5",
    genome = "unknown",
    version = "3",
    chemistry = "Single Cell 3' v2",
    original.gem.groups = 1L,
    library.ids = "custom"
  )

  message("Done: ", gen_samples[i])
}

saveRDS(gen_barcodes_post_ed, file = here::here("data/processed/genotype/emptyDrops_filtered/rds-files/genotype_barcodes_post_ed.Rds"))
```

## Smart-seq2

emptyDrops is not applicable to plate-based data. Cells are filtered using fixed thresholds for total counts and detected features, chosen by inspecting violin plots per sample.

```{r}
# Filtering thresholds (set after visual inspection)
min_counts <- 100000
max_counts <- 500000
min_features <- 2500
max_features <- 10500

smart_barcodes_post_filtering <- list()
smart_filtered_matrices <- list()

for (i in seq_along(smart_count_matrices)) {
  total_counts <- Matrix::colSums(smart_count_matrices[[i]])
  detected_features <- Matrix::colSums(smart_count_matrices[[i]] > 0)

  keep_cells <- (total_counts >= min_counts) &
                (total_counts <= max_counts) &
                (detected_features >= min_features) &
                (detected_features <= max_features)

  smart_barcodes_post_filtering[[i]] <- colnames(smart_count_matrices[[i]])[keep_cells]
  names(smart_barcodes_post_filtering)[i] <- names(smart_count_matrices)[i]

  saveRDS(smart_barcodes_post_filtering[[i]], here::here(paste0(
    "data/processed/smart-seq2/filtered/rds-files/",
    names(smart_count_matrices)[i], "_barcodes_post_filtering.Rds"
  )))

  mat <- smart_count_matrices[[i]][, keep_cells]
  mat <- mat[rowSums(mat) > 0, ]
  smart_filtered_matrices[[i]] <- mat

  write10xCounts(
    here::here(paste0(
      "data/processed/smart-seq2/filtered/h5-files/",
      names(smart_count_matrices)[i], "_post_filtering.h5"
    )),
    mat,
    barcodes = colnames(mat),
    gene.id = rownames(mat),
    gene.symbol = rownames(mat),
    gene.type = "Gene Expression",
    overwrite = TRUE,
    type = "HDF5",
    genome = "unknown",
    version = "3",
    chemistry = "Smart-seq2",
    original.gem.groups = 1L,
    library.ids = "custom"
  )

  message("Done: ", names(smart_count_matrices)[i])
}

saveRDS(smart_barcodes_post_filtering, file = here::here("data/processed/smart-seq2/filtered/rds-files/smart_barcodes_post_filtering.Rds"))
```
