---
title: "06_further_preprocessing"
format: pdf
editor: visual
---

Post-decontamination preprocessing: feature name correction, mitochondrial filtering, metadata addition, species-of-origin labelling, dimensionality reduction, and clustering.

------------------------------------------------------------------------

# Load Seurat objects

```{r}
lc_seu <- readRDS(here::here("data/processed/low_complexity/seurat_object/lc_seu_before_processing.Rds"))
mc_human_seu <- readRDS(here::here("data/processed/medium_complexity/seurat_object/mc_human_seu_before_processing.Rds"))
mc_mouse_seu <- readRDS(here::here("data/processed/medium_complexity/seurat_object/mc_mouse_seu_before_processing.Rds"))
hc_seu <- readRDS(here::here("data/processed/high_complexity/seurat_object/hc_seu_before_processing.Rds"))
gen_seu <- readRDS(here::here("data/processed/genotype/seurat_object/gen_seu_before_processing.Rds"))
smart_seu <- readRDS(here::here("data/processed/smart-seq2/seurat_object/smart_seu_before_processing.Rds"))
synt_seu <- readRDS(here::here("data/processed/synthetic/seurat_object/synt_seu_before_processing.Rds"))
synt_seu_wo_ambient <- readRDS(here::here("data/processed/synthetic/seurat_object/synt_seu_wo_ambient_before_processing.Rds"))
```

------------------------------------------------------------------------

# Correct feature names

Standardises gene name formatting across all assays. Not needed for the genotype dataset (names are already consistent).

```{r}
lc_seu <- correct_rownames(lc_seu)
mc_human_seu <- correct_rownames(mc_human_seu)
mc_mouse_seu <- correct_rownames(mc_mouse_seu)
hc_seu <- correct_rownames(hc_seu)
```

### Synthetic — convert Ensembl IDs to gene symbols

The synthetic dataset uses Ensembl IDs as feature names. These are converted to HGNC gene symbols using a local annotation database (org.Hs.eg.db) to avoid network-dependent BioMart queries.

```{r}
synt_seu <- convert_ensembl_ids_to_gene_names(synt_seu)
synt_seu_wo_ambient <- convert_ensembl_ids_to_gene_names(synt_seu_wo_ambient)
```

After conversion, rownames across all assays are synchronised to match the RNA assay. This is required because Ensembl-to-symbol mapping may produce different results for each assay if applied independently.

```{r}
# Synchronise rownames of all non-RNA assays to the RNA assay
for (seu_obj in c("synt_seu", "synt_seu_wo_ambient")) {
  obj <- get(seu_obj)
  assays <- names(obj@assays)
  for (assay in assays[-1]) {
    counts <- GetAssayData(obj, assay = assay, layer = "counts")
    rownames(counts) <- rownames(obj@assays$RNA)
    obj[[assay]] <- CreateAssayObject(counts = counts)
  }
  assign(seu_obj, obj)
}
rm(obj, counts)
```

------------------------------------------------------------------------

# Remove cells with high mitochondrial content

Thresholds were chosen per dataset after visual inspection of violin plots.

| Dataset           | MT pattern            | Threshold |
|-------------------|-----------------------|-----------|
| Low complexity    | mm10-mt- / GRCh38-MT- | 15 %      |
| Medium complexity | mm10-mt- / GRCh38-MT- | 15 %      |
| High complexity   | mm10-mt- / GRCh38-MT- | 1 %       |
| Genotype          | mt-                   | 60 %      |
| Smart-seq2        | MT-                   | 6 %       |
| Synthetic         | MT-                   | 10 %      |

```{r}
lc_seu <- remove_mito_genes(lc_seu, mito_pattern = c("mm10-mt-", "GRCh38-MT-"), mt_threshold = 0.15)
mc_human_seu <- remove_mito_genes(mc_human_seu, mito_pattern = c("mm10-mt-", "GRCh38-MT-"), mt_threshold = 0.15)
mc_mouse_seu <- remove_mito_genes(mc_mouse_seu, mito_pattern = c("mm10-mt-", "GRCh38-MT-"), mt_threshold = 0.15)
hc_seu <- remove_mito_genes(hc_seu, mito_pattern = c("mm10-mt-", "GRCh38-MT-"), mt_threshold = 0.01)
gen_seu <- remove_mito_genes(gen_seu, mito_pattern = "mt-",                        mt_threshold = 0.60)
smart_seu <- remove_mito_genes(smart_seu, mito_pattern = "MT-",                        mt_threshold = 0.06)
synt_seu <- remove_mito_genes(synt_seu, mito_pattern = "MT-",                        mt_threshold = 0.10)
synt_seu_wo_ambient <- remove_mito_genes(synt_seu_wo_ambient, mito_pattern = "MT-", mt_threshold = 0.10)
```

------------------------------------------------------------------------

# Add metadata

Only needed for the genotype and synthetic datasets.

## Genotype

```{r}
cell_metadata <- readRDS(here::here("data/processed/genotype/metadata/cell_metadata.RDS"))
cell_metadata$barcodes <- paste(cell_metadata$cell, cell_metadata$replicate, sep = "_")
rownames(cell_metadata) <- cell_metadata$barcodes
cell_metadata[, c("replicate", "cell", "barcodes")] <- NULL

gen_seu <- AddMetaData(gen_seu, metadata = cell_metadata)
```

## Synthetic

```{r}
synt_metadata <- readRDS(here::here("data/raw/synthetic/metadata/synt_metadata.Rds"))
synt_seu <- AddMetaData(synt_seu, metadata = synt_metadata)

# For the ambient-free object, only the celltype column is needed
labels <- synt_metadata[match(colnames(synt_seu_wo_ambient), rownames(synt_metadata)), ]
synt_seu_wo_ambient$celltype <- labels$celltype
```

------------------------------------------------------------------------

# Label cells by species of origin

Adds an `origin` metadata column (human / mouse) based on the ratio of species-specific UMIs. Only needed for the barnyard and species-mixing datasets.

```{r}
lc_seu <- determine_species_origin(lc_seu)
mc_human_seu <- determine_species_origin(mc_human_seu)
mc_mouse_seu <- determine_species_origin(mc_mouse_seu)
hc_seu <- determine_species_origin(hc_seu)
```

------------------------------------------------------------------------

# Dimensionality reduction and clustering

Parameters were chosen interactively from clustree and UMAP visualisations and are stored in `seu@misc` for reference.

## Low complexity

Resolution yields the two true clusters (human and mouse cell lines).

```{r}
lc_seu <- NormalizeData(lc_seu, verbose = FALSE)
lc_seu <- FindVariableFeatures(lc_seu, verbose = FALSE)
lc_seu <- ScaleData(lc_seu, verbose = FALSE)
lc_seu <- RunPCA(lc_seu, verbose = FALSE)
lc_seu <- RunUMAP(lc_seu, dims = 1:10, verbose = FALSE)
lc_seu <- FindNeighbors(lc_seu, dims = 1:10, verbose = FALSE)
lc_seu <- FindClusters(lc_seu, resolution = 0.03, verbose = FALSE)

lc_seu@misc$dims_used_further_preprocessing <- 10
lc_seu@misc$res_used_further_preprocessing  <- 0.03
```

## Medium complexity — human samples

Harmony integration for batch correction across donors.

```{r}
mc_human_seu <- NormalizeData(mc_human_seu, verbose = FALSE)
mc_human_seu <- FindVariableFeatures(mc_human_seu, verbose = FALSE)
mc_human_seu <- ScaleData(mc_human_seu, verbose = FALSE)
mc_human_seu <- RunPCA(mc_human_seu, verbose = FALSE)
mc_human_seu <- RunHarmony(mc_human_seu, group.by.vars = "orig.ident", verbose = FALSE)
mc_human_seu <- RunUMAP(mc_human_seu, dims = 1:12, reduction = "harmony", verbose = FALSE)
mc_human_seu <- FindNeighbors(mc_human_seu, dims = 1:12, reduction = "harmony", verbose = FALSE)
mc_human_seu <- FindClusters(mc_human_seu, resolution = 0.2, verbose = FALSE)

mc_human_seu@misc$dims_used_further_preprocessing <- 12
mc_human_seu@misc$res_used_further_preprocessing <- 0.2
```

## Medium complexity — mouse samples

```{r}
mc_mouse_seu <- NormalizeData(mc_mouse_seu, verbose = FALSE)
mc_mouse_seu <- FindVariableFeatures(mc_mouse_seu, verbose = FALSE)
mc_mouse_seu <- ScaleData(mc_mouse_seu, verbose = FALSE)
mc_mouse_seu <- RunPCA(mc_mouse_seu, verbose = FALSE)
mc_mouse_seu <- RunHarmony(mc_mouse_seu, group.by.vars = "orig.ident", verbose = FALSE)
mc_mouse_seu <- RunUMAP(mc_mouse_seu, dims = 1:10, reduction = "harmony", verbose = FALSE)
mc_mouse_seu <- FindNeighbors(mc_mouse_seu, dims = 1:10, reduction = "harmony", verbose = FALSE)
mc_mouse_seu <- FindClusters(mc_mouse_seu, resolution = 0.2, verbose = FALSE)

mc_mouse_seu@misc$dims_used_further_preprocessing <- 10
mc_mouse_seu@misc$res_used_further_preprocessing <- 0.2
```

## High complexity

```{r}
hc_seu <- NormalizeData(hc_seu, verbose = FALSE)
hc_seu <- FindVariableFeatures(hc_seu, verbose = FALSE)
hc_seu <- ScaleData(hc_seu, verbose = FALSE)
hc_seu <- RunPCA(hc_seu, verbose = FALSE)
hc_seu <- RunUMAP(hc_seu, dims = 1:15, verbose = FALSE)
hc_seu <- FindNeighbors(hc_seu, dims = 1:15, verbose = FALSE)
hc_seu <- FindClusters(hc_seu, resolution = 0.4, verbose = FALSE)

hc_seu@misc$dims_used_further_preprocessing <- 15
hc_seu@misc$res_used_further_preprocessing <- 0.4
```

## Genotype

```{r}
gen_seu <- NormalizeData(gen_seu, verbose = FALSE)
gen_seu <- FindVariableFeatures(gen_seu, verbose = FALSE)
gen_seu <- ScaleData(gen_seu, verbose = FALSE)
gen_seu <- RunPCA(gen_seu, verbose = FALSE)
gen_seu <- RunUMAP(gen_seu, dims = 1:15, verbose = FALSE)
gen_seu <- FindNeighbors(gen_seu, dims = 1:15, verbose = FALSE)
gen_seu <- FindClusters(gen_seu, resolution = 0.15, verbose = FALSE)

gen_seu@misc$dims_used_further_preprocessing <- 15
gen_seu@misc$res_used_further_preprocessing <- 0.15
```

## Smart-seq2

```{r}
smart_seu <- NormalizeData(smart_seu, verbose = FALSE)
smart_seu <- FindVariableFeatures(smart_seu, verbose = FALSE)
smart_seu <- ScaleData(smart_seu, verbose = FALSE)
smart_seu <- RunPCA(smart_seu, verbose = FALSE)
smart_seu <- RunUMAP(smart_seu, dims = 1:10, verbose = FALSE)
smart_seu <- FindNeighbors(smart_seu, dims = 1:10, verbose = FALSE)
smart_seu <- FindClusters(smart_seu, resolution = 0.5, verbose = FALSE)

smart_seu@misc$dims_used_further_preprocessing <- 10
smart_seu@misc$res_used_further_preprocessing <- 0.5
```

## Synthetic — with ambient RNA

```{r}
synt_seu <- NormalizeData(synt_seu, verbose = FALSE)
synt_seu <- FindVariableFeatures(synt_seu, verbose = FALSE)
synt_seu <- ScaleData(synt_seu, verbose = FALSE)
synt_seu <- RunPCA(synt_seu, verbose = FALSE)
synt_seu <- RunUMAP(synt_seu, dims = 1:4, verbose = FALSE)
synt_seu <- FindNeighbors(synt_seu, dims = 1:4, verbose = FALSE)
synt_seu <- FindClusters(synt_seu, resolution = 0.05, verbose = FALSE)

synt_seu@misc$dims_used_further_preprocessing <- 4
synt_seu@misc$res_used_further_preprocessing <- 0.05
```

## Synthetic — without ambient RNA

```{r}
synt_seu_wo_ambient <- NormalizeData(synt_seu_wo_ambient, verbose = FALSE)
synt_seu_wo_ambient <- FindVariableFeatures(synt_seu_wo_ambient, verbose = FALSE)
synt_seu_wo_ambient <- ScaleData(synt_seu_wo_ambient, verbose = FALSE)
synt_seu_wo_ambient <- RunPCA(synt_seu_wo_ambient, verbose = FALSE)
synt_seu_wo_ambient <- RunUMAP(synt_seu_wo_ambient, dims = 1:4, verbose = FALSE)
synt_seu_wo_ambient <- FindNeighbors(synt_seu_wo_ambient, dims = 1:4, verbose = FALSE)
synt_seu_wo_ambient <- FindClusters(synt_seu_wo_ambient, resolution = 0.03, verbose = FALSE)

synt_seu_wo_ambient@misc$dims_used_further_preprocessing <- 4
synt_seu_wo_ambient@misc$res_used_further_preprocessing <- 0.03
```

------------------------------------------------------------------------

# Save final Seurat objects

```{r}
saveRDS(lc_seu, here::here("data/processed/low_complexity/seurat_object/lc_seu.Rds"))
saveRDS(mc_human_seu, here::here("data/processed/medium_complexity/seurat_object/mc_human_seu.Rds"))
saveRDS(mc_mouse_seu, here::here("data/processed/medium_complexity/seurat_object/mc_mouse_seu.Rds"))
saveRDS(hc_seu, here::here("data/processed/high_complexity/seurat_object/hc_seu.Rds"))
saveRDS(gen_seu, here::here("data/processed/genotype/seurat_object/gen_seu.Rds"))
saveRDS(smart_seu, here::here("data/processed/smart-seq2/seurat_object/smart_seu.Rds"))
saveRDS(synt_seu, here::here("data/processed/synthetic/seurat_object/synt_seu.Rds"))
saveRDS(synt_seu_wo_ambient, here::here("data/processed/synthetic/seurat_object/synt_seu_wo_ambient.Rds"))
```
