---
title: "07_analysis_and_figures"
format: pdf
editor: visual
editor_options:
  chunk_output_type: console
---

All analysis and figure generation for the study. Outputs are saved to `results/main_figures/`, `results/supplemental_figures/`, and `results/supplemental_tables/`.

------------------------------------------------------------------------

# Load Seurat objects

```{r}
lc_seu <- readRDS(here::here("data/processed/low_complexity/seurat_object/lc_seu.Rds"))
hc_seu <- readRDS(here::here("data/processed/high_complexity/seurat_object/hc_seu.Rds"))
gen_seu <- readRDS(here::here("data/processed/genotype/seurat_object/gen_seu.Rds"))
smart_seu <- readRDS(here::here("data/processed/smart-seq2/seurat_object/smart_seu.Rds"))
synt_seu <- readRDS(here::here("data/processed/synthetic/seurat_object/synt_seu.Rds"))
synt_wo_ambient_seu <- readRDS(here::here("data/processed/synthetic/seurat_object/synt_seu_wo_ambient.Rds"))

# Medium complexity: subset to spike-in cells of the opposite species and
# remove samples with < 10 cells after subsetting
mc_human_seu <- readRDS(here::here("data/processed/medium_complexity/seurat_object/mc_human_seu.Rds"))
mc_human_seu <- subset(mc_human_seu, subset = origin == "mouse")
mc_human_seu <- subset(mc_human_seu, subset = !(orig.ident %in% names(table(mc_human_seu$orig.ident)[table(mc_human_seu$orig.ident) < 10])))
# Removed: SRR10751485, SRR10751486, SRR10751492, SRR10751494, SRR10751496
mc_human_samples <- mc_human_samples[!mc_human_samples %in% c("SRR10751485", "SRR10751486", "SRR10751492", "SRR10751494", "SRR10751496")]
saveRDS(mc_human_seu, here::here("data/processed/medium_complexity/seurat_object/mc_human_seu_subset.Rds"))

mc_mouse_seu <- readRDS(here::here("data/processed/medium_complexity/seurat_object/mc_mouse_seu.Rds"))
mc_mouse_seu <- subset(mc_mouse_seu, subset = origin == "human")
mc_mouse_seu <- subset(mc_mouse_seu, subset = !(orig.ident %in% names(table(mc_mouse_seu$orig.ident)[table(mc_mouse_seu$orig.ident) < 10])))
# Removed: SRR10751508, SRR10751514
mc_mouse_samples <- mc_mouse_samples[!mc_mouse_samples %in% c("SRR10751508", "SRR10751514")]
saveRDS(mc_mouse_seu, here::here("data/processed/medium_complexity/seurat_object/mc_mouse_seu_subset.Rds"))
```

------------------------------------------------------------------------

# Characterising ambient RNA

Ambient characterisation is performed for all droplet-based datasets (low, medium, high complexity, and genotype). Smart-seq2 and synthetic datasets are excluded as they lack a meaningful empty droplet pool.

## Divide counts into ambient and endogenous compartments

### Low complexity

The low complexity dataset is a barnyard experiment and can be evaluated in both directions: mouse genes contaminating human cells, and vice versa.

```{r}
# Human cells contaminated by mouse genes
lc_raw_human <- extract_raw_counts(seu = lc_seu, origin = "human")
lc_raw_human_ambient <- extract_ambient_counts(lc_raw_human,    ambient_pattern    = "^mm10-")
lc_raw_human_endogenous <- extract_endogenous_counts(lc_raw_human, endogenous_pattern = "^GRCh38-")

# Mouse cells contaminated by human genes
lc_raw_mouse <- extract_raw_counts(seu = lc_seu, origin = "mouse")
lc_raw_mouse_ambient <- extract_ambient_counts(lc_raw_mouse,    ambient_pattern    = "^GRCh38-")
lc_raw_mouse_endogenous <- extract_endogenous_counts(lc_raw_mouse, endogenous_pattern = "^mm10-")
```

### Medium complexity

Only spike-in cells (cells of the opposite species from the sample) are used to evaluate decontamination.

```{r}
# Human samples: mouse spike-in cells carry ambient human (GRCh38) signal
mc_raw_human <- extract_raw_counts(seu = mc_human_seu, origin = "mouse")
mc_raw_human_ambient <- extract_ambient_counts(mc_raw_human, ambient_pattern = "^GRCh38-")
mc_raw_human_endogenous <- extract_endogenous_counts(mc_raw_human, endogenous_pattern = "^mm10-")

mc_raw_human_not_spikein <- extract_raw_counts(seu = mc_human_seu, origin = "human")
mc_raw_human_ambient_not_spikein <- extract_ambient_counts(mc_raw_human_not_spikein, ambient_pattern = "^mm10-")
mc_raw_human_endogenous_not_spikein <- extract_endogenous_counts(mc_raw_human_not_spikein, endogenous_pattern = "^GRCh38-")

# Mouse samples: human spike-in cells carry ambient mouse (mm10) signal
mc_raw_mouse <- extract_raw_counts(seu = mc_mouse_seu, origin = "human")
mc_raw_mouse_ambient <- extract_ambient_counts(mc_raw_mouse, ambient_pattern = "^mm10-")
mc_raw_mouse_endogenous <- extract_endogenous_counts(mc_raw_mouse, endogenous_pattern = "^GRCh38-")

mc_raw_mouse_not_spikein <- extract_raw_counts(seu = mc_mouse_seu, origin = "mouse")
mc_raw_mouse_ambient_not_spikein <- extract_ambient_counts(mc_raw_mouse_not_spikein,  ambient_pattern = "^GRCh38-")
mc_raw_mouse_endogenous_not_spikein <- extract_endogenous_counts(mc_raw_mouse_not_spikein, endogenous_pattern = "^mm10-")
```

### High complexity

```{r}
hc_raw_human <- extract_raw_counts(seu = hc_seu, origin = "human")
hc_raw_human_ambient <- extract_ambient_counts(hc_raw_human, ambient_pattern = "^mm10-")
hc_raw_human_endogenous <- extract_endogenous_counts(hc_raw_human, endogenous_pattern = "^GRCh38-")

hc_raw_mouse <- extract_raw_counts(seu = hc_seu, origin = "mouse")
hc_raw_mouse_ambient <- extract_ambient_counts(hc_raw_mouse, ambient_pattern = "^GRCh38-")
hc_raw_mouse_endogenous <- extract_endogenous_counts(hc_raw_mouse, endogenous_pattern = "^mm10-")
```

### Genotype

For the genotype dataset, ambient fractions are estimated from SNP data using `calc_ambient_from_SNPs()` (defined in `02_functions.qmd`). The genotype_estimation folder comes from the study from which the dataset was retrieved.

Each SNP position is also classified as exonic using a BED intersection (see `03_download_datasets.qmd` for the exon BED file preparation).

```{r}
# Calculate ambient fractions per gene per cell from SNP calls
csc_all <- dplyr::bind_rows(lapply(gen_samples, calc_ambient_from_SNPs))
```

```{r}
# Classify SNP positions as exonic or intronic
exonic <- read.table(
  here::here("data/raw/reference_sources/refdata-gex-GRCm39-2024-A/genes/variants_in_exons.bed"),
  header = FALSE
) |>
  setNames(c("chr", "start", "end")) |>
  dplyr::mutate(pos = start + 1) |>
  dplyr::select(chr, pos)

csc_all <- csc_all |>
  tidyr::separate(Position, into = c("chr", "pos"), convert = TRUE) |>
  dplyr::mutate(
    chr       = paste0("chr", chr),
    is_exonic = (chr %in% exonic$chr & pos %in% exonic$pos)
  )

saveRDS(csc_all, here::here("data/processed/genotype/genotype_estimation/all/csc_all_with_exon_information.Rds"))
```

```{r}
# Build ambient and endogenous matrices from SNP-derived fractions
gen_raw_counts <- as.matrix(gen_seu@assays$RNA$counts)
gen_raw_ambient <- matrix(NA, nrow = nrow(gen_raw_counts), ncol = ncol(gen_raw_counts), dimnames = dimnames(gen_raw_counts))

csc_filt <- csc_all[csc_all$Symbol %in% rownames(gen_raw_ambient) & csc_all$Barcode %in% colnames(gen_raw_ambient), ]

row_idx <- match(csc_filt$Symbol, rownames(gen_raw_ambient))
col_idx <- match(csc_filt$Barcode, colnames(gen_raw_ambient))
gen_raw_ambient[cbind(row_idx, col_idx)] <- gen_raw_counts[cbind(row_idx, col_idx)] * csc_filt$Ambient

gen_raw_endogenous <- as.matrix(gen_raw_counts - gen_raw_ambient)
```

------------------------------------------------------------------------

## Ambient load per cell

### Low complexity

```{r}
lc_ambient_load_percentages <- calc_ambient_load_percentages(
  ambient_list = list(human = lc_raw_human_ambient, mouse = lc_raw_mouse_ambient),
  raw_list = list(human = lc_raw_human, mouse = lc_raw_mouse)
)

lc_avg_ambient_load_percentage <- calc_avg_ambient_load_percentage(lc_ambient_load_percentages)

lc_amb_pct_hist <- make_histogram_of_ambient_load(
  ambient_load = lc_ambient_load_percentages,
  avg_ambient_load = lc_avg_ambient_load_percentage
)
saveRDS(lc_amb_pct_hist, here::here("results/main_figures/figure_2/lc_amb_pct_hist.rds"))
```

### Medium complexity — human samples

```{r}
mc_human_samples_current <- unique(mc_human_seu$orig.ident)
mc_ambient_load_percentages_human <- calc_ambient_load_percentages(
  ambient_list = list(human = mc_raw_human_ambient),
  raw_list = list(human = mc_raw_human),
  seu = mc_human_seu,
  samples = mc_human_samples_current
)
colnames(mc_ambient_load_percentages_human)[5] <- "value"

mc_avg_ambient_load_human <- calc_avg_ambient_load_percentage(mc_ambient_load_percentages_human)
mc_avg_ambient_load_human$group <- factor(mc_avg_ambient_load_human$group,
                                           levels = rev(paste0("S", seq_len(14))))
```

### Medium complexity — mouse samples

```{r}
mc_mouse_samples_current <- unique(mc_mouse_seu$orig.ident)
mc_ambient_load_percentages_mouse <- calc_ambient_load_percentages(
  ambient_list = list(mouse = mc_raw_mouse_ambient),
  raw_list = list(mouse = mc_raw_mouse),
  seu = mc_mouse_seu,
  samples = mc_mouse_samples_current
)
colnames(mc_ambient_load_percentages_mouse)[5] <- "value"

mc_avg_ambient_load_mouse <- calc_avg_ambient_load_percentage(mc_ambient_load_percentages_mouse)
mc_avg_ambient_load_mouse$group <- factor(mc_avg_ambient_load_mouse$group,
                                           levels = rev(paste0("S", seq_len(13))))
```

### Medium complexity — combined plot

```{r}
mc_ambient_load_percentages_combined <- rbind(
  mc_ambient_load_percentages_human,
  mc_ambient_load_percentages_mouse
)
mc_avg_ambient_load_combined <- rbind(mc_avg_ambient_load_human, mc_avg_ambient_load_mouse)
mc_ambient_load_percentages_combined$group <- factor(
  mc_ambient_load_percentages_combined$group, levels = rev(paste0("S", seq_len(27)))
)

mc_amb_pct_ridge <- ambient_load_ridge_plot(
  ambient_load = mc_ambient_load_percentages_combined,
  avg_ambient_load = mc_avg_ambient_load_combined
)
saveRDS(mc_amb_pct_ridge, here::here("results/main_figures/figure_2/mc_amb_pct_ridge.rds"))
```

### High complexity

```{r}
hc_ambient_load_percentages <- calc_ambient_load_percentages(
  ambient_list = list(human = hc_raw_human_ambient, mouse = hc_raw_mouse_ambient),
  raw_list = list(human = hc_raw_human,         mouse = hc_raw_mouse),
  seu = hc_seu,
  samples = hc_samples
)
hc_avg_ambient_load <- calc_avg_ambient_load_percentage(hc_ambient_load_percentages)

hc_ambient_load_percentages$group <- factor(hc_ambient_load_percentages$group, levels = rev(paste0("S", seq_len(16))))
hc_amb_pct_ridge <- ambient_load_ridge_plot(hc_ambient_load_percentages, hc_avg_ambient_load)
saveRDS(hc_amb_pct_ridge, here::here("results/main_figures/figure_2/hc_amb_pct_ridge.rds"))
```

### Genotype

```{r}
gen_ambient_load_percentages <- calc_ambient_load_percentages(
  ambient_list = list(CAST = gen_raw_ambient),
  raw_list = list(CAST = gen_raw_counts),
  seu = gen_seu,
  samples = gen_samples
)
gen_avg_ambient_load <- calc_avg_ambient_load_percentage(gen_ambient_load_percentages)

gen_ambient_load_percentages$group <- factor(gen_ambient_load_percentages$group, levels = rev(paste0("S", seq_len(5))))
gen_amb_pct_ridge <- ambient_load_ridge_plot(gen_ambient_load_percentages, gen_avg_ambient_load)
saveRDS(gen_amb_pct_ridge, here::here("results/main_figures/figure_2/gen_amb_pct_ridge.rds"))
```

### Summary statistics: QCD and strongly affected barcodes

```{r}
# Combine all datasets
ambient_load_all_datasets <- dplyr::bind_rows(
  dplyr::mutate(lc_ambient_load_percentages, dataset = "low_complexity"),
  dplyr::mutate(mc_ambient_load_percentages_combined, dataset = "medium_complexity"),
  dplyr::mutate(hc_ambient_load_percentages, dataset = "high_complexity"),
  dplyr::mutate(gen_ambient_load_percentages, dataset = "genotype")
)
write_xlsx(ambient_load_all_datasets, path = here::here("results/supplemental_tables/ambient_load_per_cell.xlsx"))

# QCD within samples
qcd_within_samples <- ambient_load_all_datasets |>
  dplyr::group_by(dataset, sample) |>
  dplyr::summarise(QCD_within_sample = calc_qcd(ambient_percent), .groups = "drop")
mean(qcd_within_samples$QCD_within_sample)

# QCD between samples within a study
sample_means <- ambient_load_all_datasets |>
  dplyr::group_by(dataset, sample) |>
  dplyr::summarise(mean_value = mean(ambient_percent), .groups = "drop")

qcd_between_samples <- sample_means |>
  dplyr::group_by(dataset) |>
  dplyr::summarise(QCD_between_samples = if (dplyr::n() > 1) calc_qcd(mean_value) else NA_real_)
mean(qcd_between_samples$QCD_between_samples, na.rm = TRUE)

# QCD between studies
study_means <- sample_means |>
  dplyr::group_by(dataset) |>
  dplyr::summarise(study_mean = mean(mean_value), .groups = "drop")
calc_qcd(study_means$study_mean)

# Fraction of top-10%-UMI cells with > 10% ambient contamination
umi_cutoff <- quantile(ambient_load_all_datasets$total_umis, 0.9, na.rm = TRUE)
top10_cells <- dplyr::filter(ambient_load_all_datasets, total_umis >= umi_cutoff)
mean(top10_cells$ambient_percent >= 10, na.rm = TRUE) * 100
```

------------------------------------------------------------------------

## Variation explained by cellular parameters

Exon fraction, mitochondrial fraction, protein-coding fraction, and ribosomal fraction are used as covariates to explain variation in ambient RNA load.

### Exon fractions

#### Low complexity

```{r}
lc_exon_fraction <- calc_exon_fraction(
  path = here::here("data/raw/low_complexity/FastQ/exon_intron/Solo.out"),
  object = lc_seu,
  name = "Barnyard"
)
lc_exon_fraction_per_feature <- lc_exon_fraction$features
lc_exon_fraction_per_cell <- lc_exon_fraction$cells
```

#### Medium complexity — human samples

```{r}
mc_exon_list_human <- list()
for (samp in mc_human_samples) {
  mc_exon_list_human[[samp]] <- calc_exon_fraction(
    path   = here::here(paste0("data/raw/medium_complexity/", samp, "_STARsolo_Alignment/", samp, "_Solo.out/")),
    object = mc_human_seu,
    name = samp,
    paste1 = FALSE
  )
}
mc_exon_fraction_per_feature_human <- unlist(lapply(mc_exon_list_human, `[[`, "features"))
mc_exon_fraction_per_cell_human <- unlist(lapply(mc_exon_list_human, `[[`, "cells"))
```

#### Medium complexity — mouse samples

```{r}
mc_exon_list_mouse <- list()
for (samp in mc_mouse_samples) {
  mc_exon_list_mouse[[samp]] <- calc_exon_fraction(
    path   = here::here(paste0("data/raw/medium_complexity/", samp, "_STARsolo_Alignment/", samp, "_Solo.out/")),
    object = mc_mouse_seu,
    name = samp,
    paste1 = FALSE
  )
}
mc_exon_fraction_per_feature_mouse <- unlist(lapply(mc_exon_list_mouse, `[[`, "features"))
mc_exon_fraction_per_cell_mouse <- unlist(lapply(mc_exon_list_mouse, `[[`, "cells"))

mc_exon_fraction_per_feature_combined <- c(mc_exon_fraction_per_feature_human, mc_exon_fraction_per_feature_mouse)
mc_exon_fraction_per_cell_combined <- c(mc_exon_fraction_per_cell_human,    mc_exon_fraction_per_cell_mouse)
```

#### High complexity

```{r}
hc_exon_list <- list()
for (samp in hc_samples) {
  hc_exon_list[[samp]] <- calc_exon_fraction(
    path = here::here(paste0("data/raw/high_complexity/", samp, "_STARsolo_Alignment/", samp, "_Solo.out/")),
    object = hc_seu,
    name = samp,
    paste1 = FALSE
  )
}
hc_exon_fraction_per_feature <- unlist(lapply(hc_exon_list, `[[`, "features"))
hc_exon_fraction_per_cell <- unlist(lapply(hc_exon_list, `[[`, "cells"))
```

#### Genotype

Exon fraction is derived from the SNP call data rather than STARsolo.

```{r}
csc_exons <- csc_all[csc_all$is_exonic == TRUE, ]

gen_raw_exon <- matrix(NA, nrow = nrow(gen_raw_counts), ncol = ncol(gen_raw_counts),
                        dimnames = dimnames(gen_raw_counts))
row_idx <- match(csc_exons$Symbol,  rownames(gen_raw_counts))
col_idx <- match(csc_exons$Barcode, colnames(gen_raw_counts))
gen_raw_exon[cbind(row_idx, col_idx)] <- gen_raw_counts[cbind(row_idx, col_idx)]

gen_exon_fraction_per_cell <- colSums(gen_raw_exon, na.rm = TRUE) / colSums(gen_raw_counts, na.rm = TRUE)
gen_exon_fraction_per_feature <- rowSums(gen_raw_exon, na.rm = TRUE) / rowSums(gen_raw_counts, na.rm = TRUE)
```

### Ambient load vs exon fraction

```{r}
# Low complexity
lc_amb_load_vs_exon_fraction <- match_ambient_load_with_exon_fractions(
  exon_fraction_per_feature = lc_exon_fraction_per_feature,
  raw_human_ambient = lc_raw_human_ambient,
  raw_mouse_ambient = lc_raw_mouse_ambient
)
lc_amb_load_vs_exon_frac <- make_scatterplot_of_amb_load_vs_exon_fraction(lc_amb_load_vs_exon_fraction)
saveRDS(lc_amb_load_vs_exon_frac, here::here("results/supplemental_figures/supp_fig_2/lc_amb_load_vs_exon_frac.rds"))
lc_amb_load_vs_exon_fraction$dataset <- "low_complexity"
write_xlsx(lc_amb_load_vs_exon_fraction, path = here::here("results/supplemental_tables/lc_ambient_load_vs_exon_fraction.xlsx"))

# Medium complexity (uses the species-aware matching function)
mc_amb_load_vs_exon_fraction <- match_ambient_load_with_exon_fractions_mc(
  exon_fraction_per_feature_1 = mc_exon_fraction_per_feature_human,
  exon_fraction_per_feature_2 = mc_exon_fraction_per_feature_mouse,
  raw_human_ambient = mc_raw_human_ambient,
  raw_mouse_ambient = mc_raw_mouse_ambient
)
mc_amb_load_vs_exon_frac <- make_scatterplot_of_amb_load_vs_exon_fraction(mc_amb_load_vs_exon_fraction)
saveRDS(mc_amb_load_vs_exon_frac, here::here("results/supplemental_figures/supp_fig_2/mc_amb_load_vs_exon_frac.rds"))
mc_amb_load_vs_exon_fraction$dataset <- "medium_complexity"
write_xlsx(mc_amb_load_vs_exon_fraction, path = here::here("results/supplemental_tables/mc_ambient_load_vs_exon_fraction.xlsx"))

# High complexity
hc_amb_load_vs_exon_fraction <- match_ambient_load_with_exon_fractions(
  exon_fraction_per_feature = hc_exon_fraction_per_feature,
  raw_human_ambient = hc_raw_human_ambient,
  raw_mouse_ambient = hc_raw_mouse_ambient
)
hc_amb_load_vs_exon_frac <- make_scatterplot_of_amb_load_vs_exon_fraction(hc_amb_load_vs_exon_fraction)
saveRDS(hc_amb_load_vs_exon_frac, here::here("results/supplemental_figures/supp_fig_2/hc_amb_load_vs_exon_frac.rds"))
hc_amb_load_vs_exon_fraction$dataset <- "high_complexity"
write_xlsx(hc_amb_load_vs_exon_fraction, path = here::here("results/supplemental_tables/hc_ambient_load_vs_exon_fraction.xlsx"))

# Genotype
gen_ambient_load_vs_exon_fraction <- data.frame(
  exon_fraction = gen_exon_fraction_per_feature,
  log_ambient_expression = log1p(rowSums(gen_raw_ambient, na.rm = TRUE))
)
gen_amb_load_vs_exon_frac <- make_scatterplot_of_amb_load_vs_exon_fraction(gen_ambient_load_vs_exon_fraction)
saveRDS(gen_amb_load_vs_exon_frac, here::here("results/supplemental_figures/supp_fig_2/gen_amb_load_vs_exon_frac.rds"))
gen_ambient_load_vs_exon_fraction$dataset <- "genotype"
write_xlsx(gen_ambient_load_vs_exon_fraction, path = here::here("results/supplemental_tables/gen_ambient_load_vs_exon_fraction.xlsx"))

rm(lc_amb_load_vs_exon_frac, mc_amb_load_vs_exon_frac, hc_amb_load_vs_exon_frac, gen_amb_load_vs_exon_frac)
```

#### Summary: percentage of ambient UMIs from top-10%-exon-fraction genes

```{r}
for (ds in c("lc", "mc", "hc", "gen")) {
  df_name  <- paste0(ds, if (ds == "gen") "_ambient_load_vs_exon_fraction" else "_amb_load_vs_exon_fraction")
  df <- get(df_name)

  threshold <- quantile(df$exon_fraction, 0.90, na.rm = TRUE)
  top10 <- df[!is.na(df$exon_fraction) & df$exon_fraction >= threshold, ]
  above90 <- df[!is.na(df$exon_fraction) & df$exon_fraction > 0.9, ]

  pct_top10 <- sum(expm1(top10$log_ambient_expression),  na.rm = TRUE) /
                sum(expm1(df$log_ambient_expression),     na.rm = TRUE) * 100
  pct_above90 <- sum(expm1(above90$log_ambient_expression), na.rm = TRUE) /
                 sum(expm1(df$log_ambient_expression),      na.rm = TRUE) * 100

  message(ds, " — % ambient from top-10% exon fraction genes: ", round(pct_top10, 1))
  message(ds, " — % ambient from genes with exon fraction > 0.9: ", round(pct_above90, 1))
}
```

### Gene biotypes

```{r}
# Low complexity
lc_gene_info <- get_gene_biotypes(
  path = here::here("data/raw/low_complexity/FastQ/GRCh38_and_mm10-2020-A_build/reference/geneInfo.tab"),
  seu  = lc_seu
)
lc_pc_fraction_per_cell <- calc_fraction_pc_genes_per_cell(lc_seu, lc_gene_info)
lc_ribo_fraction_per_cell <- calc_fraction_ribo_genes_per_cell(lc_seu)
lc_mt_fraction_per_cell <- lc_seu$fraction.mt

# Medium complexity — human
mc_gene_info_human <- get_gene_biotypes(
  path = here::here("data/raw/medium_complexity/FastQ/GRCh38-and-mm10-2020-A-10X-Download/star/geneInfo.tab"),
  seu  = mc_human_seu
)
mc_pc_fraction_per_cell_human <- calc_fraction_pc_genes_per_cell(mc_human_seu, mc_gene_info_human)
mc_ribo_fraction_per_cell_human <- calc_fraction_ribo_genes_per_cell(mc_human_seu)
mc_mt_fraction_per_cell_human <- mc_human_seu$fraction.mt

# Medium complexity — mouse
mc_gene_info_mouse <- get_gene_biotypes(
  path = here::here("data/raw/medium_complexity/FastQ/GRCh38-and-mm10-2020-A-10X-Download/star/geneInfo.tab"),
  seu  = mc_mouse_seu
)
mc_pc_fraction_per_cell_mouse <- calc_fraction_pc_genes_per_cell(mc_mouse_seu, mc_gene_info_mouse)
mc_ribo_fraction_per_cell_mouse <- calc_fraction_ribo_genes_per_cell(mc_mouse_seu)
mc_mt_fraction_per_cell_mouse <- mc_mouse_seu$fraction.mt

mc_pc_fraction_per_cell_combined <- c(mc_pc_fraction_per_cell_human,   mc_pc_fraction_per_cell_mouse)
mc_ribo_fraction_per_cell_combined <- c(mc_ribo_fraction_per_cell_human, mc_ribo_fraction_per_cell_mouse)
mc_mt_fraction_per_cell_combined <- c(mc_mt_fraction_per_cell_human,   mc_mt_fraction_per_cell_mouse)

# High complexity
hc_gene_info <- get_gene_biotypes(
  path = here::here("data/raw/reference_sources/GRCh38-and-mm10-2020-A-10X-Download/star/geneInfo.tab"),
  seu  = hc_seu
)
hc_pc_fraction_per_cell <- calc_fraction_pc_genes_per_cell(hc_seu, hc_gene_info)
hc_ribo_fraction_per_cell <- calc_fraction_ribo_genes_per_cell(hc_seu)
hc_mt_fraction_per_cell <- hc_seu$fraction.mt

# Genotype
gen_gene_info <- get_gene_biotypes(
  path = here::here("data/raw/reference_sources/refdata-gex-GRCm39-2024-A/star/geneInfo.tab"),
  seu  = gen_seu
)
gen_pc_fraction_per_cell <- calc_fraction_pc_genes_per_cell(gen_seu, gen_gene_info)
gen_ribo_fraction_per_cell <- calc_fraction_ribo_genes_per_cell(gen_seu)
gen_mt_fraction_per_cell <- gen_seu$fraction.mt
```

### Variance explained by cellular parameters

```{r}
# Helper: match ambient load fractions with a gene biotype covariate
make_biotype_df <- function(ambient_load_fractions, gene_biotype) {
  match_ambient_load_with_gene_biotypes(
    ambient_load_fractions = ambient_load_fractions,
    gene_biotype = gene_biotype
  )
}

# Low complexity
lc_ambient_load_fractions <- calc_ambient_load_fractions(
  raw_ambient_1 = lc_raw_mouse_ambient, raw_1 = lc_raw_mouse,
  raw_ambient_2 = lc_raw_human_ambient, raw_2 = lc_raw_human
)
lc_var_expl_human <- calc_var_explained_by_cellular_params(
  seu = lc_seu, species = "human",
  ambient_fraction = lc_ambient_load_fractions,
  ambient_exon_fraction_per_cell = make_biotype_df(lc_ambient_load_fractions, lc_exon_fraction_per_cell),
  ambient_pc_fraction_per_cell = make_biotype_df(lc_ambient_load_fractions, lc_pc_fraction_per_cell),
  ambient_ribo_fraction_per_cell = make_biotype_df(lc_ambient_load_fractions, lc_ribo_fraction_per_cell),
  ambient_mt_fraction_per_cell = make_biotype_df(lc_ambient_load_fractions, lc_mt_fraction_per_cell),
  raw_counts = lc_raw_human, ambient_counts = lc_raw_human_ambient, endogenous_counts = lc_raw_human_endogenous
)
lc_var_expl_mouse <- calc_var_explained_by_cellular_params(
  seu = lc_seu, species = "mouse",
  ambient_fraction = lc_ambient_load_fractions,
  ambient_exon_fraction_per_cell = make_biotype_df(lc_ambient_load_fractions, lc_exon_fraction_per_cell),
  ambient_pc_fraction_per_cell = make_biotype_df(lc_ambient_load_fractions, lc_pc_fraction_per_cell),
  ambient_ribo_fraction_per_cell = make_biotype_df(lc_ambient_load_fractions, lc_ribo_fraction_per_cell),
  ambient_mt_fraction_per_cell = make_biotype_df(lc_ambient_load_fractions, lc_mt_fraction_per_cell),
  raw_counts = lc_raw_mouse, ambient_counts = lc_raw_mouse_ambient, endogenous_counts = lc_raw_mouse_endogenous
)
lc_var_explained <- data.frame(
  parameters = lc_var_expl_human[[1]],
  percent_explained = (lc_var_expl_human$percent_explained + lc_var_expl_mouse$percent_explained) / 2
)
lc_var_by_cellular_params <- make_barplot_of_explained_variance(lc_var_explained)
saveRDS(lc_var_by_cellular_params, here::here("results/main_figures/figure_2/lc_var_by_celular_params.rds"))
lc_var_explained$dataset <- "low_complexity"
write_xlsx(lc_var_explained, path = here::here("results/supplemental_tables/lc_var_by_cellular_parameters.xlsx"))
rm(lc_var_expl_human, lc_var_expl_mouse, lc_var_by_cellular_params)

# Medium complexity
mc_ambient_load_fractions_human <- calc_ambient_load_fractions(
  raw_ambient_1 = mc_raw_human_ambient, raw_1 = mc_raw_human
)
mc_ambient_load_fractions_mouse <- calc_ambient_load_fractions(
  raw_ambient_1 = mc_raw_mouse_ambient, raw_1 = mc_raw_mouse
)
mc_var_expl_human <- calc_var_explained_by_cellular_params(
  seu = mc_human_seu, species = "mouse",
  ambient_fraction = mc_ambient_load_fractions_human,
  ambient_exon_fraction_per_cell = make_biotype_df(mc_ambient_load_fractions_human, mc_exon_fraction_per_cell_human),
  ambient_pc_fraction_per_cell = make_biotype_df(mc_ambient_load_fractions_human, mc_pc_fraction_per_cell_human),
  ambient_ribo_fraction_per_cell = make_biotype_df(mc_ambient_load_fractions_human, mc_ribo_fraction_per_cell_human),
  ambient_mt_fraction_per_cell = make_biotype_df(mc_ambient_load_fractions_human, mc_mt_fraction_per_cell_human),
  raw_counts = mc_raw_human, ambient_counts = mc_raw_human_ambient, endogenous_counts = mc_raw_human_endogenous
)
mc_var_expl_mouse <- calc_var_explained_by_cellular_params(
  seu = mc_mouse_seu, species = "human",
  ambient_fraction = mc_ambient_load_fractions_mouse,
  ambient_exon_fraction_per_cell = make_biotype_df(mc_ambient_load_fractions_mouse, mc_exon_fraction_per_cell_mouse),
  ambient_pc_fraction_per_cell = make_biotype_df(mc_ambient_load_fractions_mouse, mc_pc_fraction_per_cell_mouse),
  ambient_ribo_fraction_per_cell = make_biotype_df(mc_ambient_load_fractions_mouse, mc_ribo_fraction_per_cell_mouse),
  ambient_mt_fraction_per_cell = make_biotype_df(mc_ambient_load_fractions_mouse, mc_mt_fraction_per_cell_mouse),
  raw_counts = mc_raw_mouse, ambient_counts = mc_raw_mouse_ambient, endogenous_counts = mc_raw_mouse_endogenous
)
mc_var_explained <- data.frame(
  parameters = mc_var_expl_human[[1]],
  percent_explained = (mc_var_expl_human$percent_explained + mc_var_expl_mouse$percent_explained) / 2
)
mc_var_by_cellular_params <- make_barplot_of_explained_variance(mc_var_explained)
saveRDS(mc_var_by_cellular_params, here::here("results/main_figures/figure_2/mc_var_by_celular_params.rds"))
mc_var_explained$dataset <- "medium_complexity"
write_xlsx(mc_var_explained, path = here::here("results/supplemental_tables/mc_var_by_cellular_parameters.xlsx"))
rm(mc_var_expl_human, mc_var_expl_mouse, mc_var_by_cellular_params)

# High complexity
hc_ambient_load_fractions <- calc_ambient_load_fractions(
  raw_ambient_1 = hc_raw_mouse_ambient, raw_1 = hc_raw_mouse,
  raw_ambient_2 = hc_raw_human_ambient, raw_2 = hc_raw_human
)
hc_var_expl_human <- calc_var_explained_by_cellular_params(
  seu = hc_seu, species = "human",
  ambient_fraction = hc_ambient_load_fractions,
  ambient_exon_fraction_per_cell = make_biotype_df(hc_ambient_load_fractions, hc_exon_fraction_per_cell),
  ambient_pc_fraction_per_cell = make_biotype_df(hc_ambient_load_fractions, hc_pc_fraction_per_cell),
  ambient_ribo_fraction_per_cell = make_biotype_df(hc_ambient_load_fractions, hc_ribo_fraction_per_cell),
  ambient_mt_fraction_per_cell = make_biotype_df(hc_ambient_load_fractions, hc_mt_fraction_per_cell),
  raw_counts = hc_raw_human, ambient_counts = hc_raw_human_ambient, endogenous_counts = hc_raw_human_endogenous
)
hc_var_expl_mouse <- calc_var_explained_by_cellular_params(
  seu = hc_seu, species = "mouse",
  ambient_fraction = hc_ambient_load_fractions,
  ambient_exon_fraction_per_cell = make_biotype_df(hc_ambient_load_fractions, hc_exon_fraction_per_cell),
  ambient_pc_fraction_per_cell = make_biotype_df(hc_ambient_load_fractions, hc_pc_fraction_per_cell),
  ambient_ribo_fraction_per_cell = make_biotype_df(hc_ambient_load_fractions, hc_ribo_fraction_per_cell),
  ambient_mt_fraction_per_cell = make_biotype_df(hc_ambient_load_fractions, hc_mt_fraction_per_cell),
  raw_counts = hc_raw_mouse, ambient_counts = hc_raw_mouse_ambient, endogenous_counts = hc_raw_mouse_endogenous
)
hc_var_explained <- data.frame(
  parameters = hc_var_expl_human[[1]],
  percent_explained = (hc_var_expl_human$percent_explained + hc_var_expl_mouse$percent_explained) / 2
)
hc_var_by_cellular_params <- make_barplot_of_explained_variance(hc_var_explained)
saveRDS(hc_var_by_cellular_params, here::here("results/main_figures/figure_2/hc_var_by_cellular_params.rds"))
hc_var_explained$dataset <- "high_complexity"
write_xlsx(hc_var_explained, path = here::here("results/supplemental_tables/hc_var_by_cellular_parameters.xlsx"))
rm(hc_var_expl_human, hc_var_expl_mouse, hc_var_by_cellular_params)

# Genotype
gen_ambient_load_fractions <- calc_ambient_load_fractions(
  raw_ambient_1 = gen_raw_ambient, raw_1 = gen_raw_counts
)
gen_var_explained <- calc_var_explained_by_cellular_params_geno(
  seu = gen_seu,
  ambient_fraction = gen_ambient_load_fractions,
  ambient_exon_fraction_per_cell = make_biotype_df(gen_ambient_load_fractions, gen_exon_fraction_per_cell),
  ambient_pc_fraction_per_cell = make_biotype_df(gen_ambient_load_fractions, gen_pc_fraction_per_cell),
  ambient_ribo_fraction_per_cell = make_biotype_df(gen_ambient_load_fractions, gen_ribo_fraction_per_cell),
  ambient_mt_fraction_per_cell = make_biotype_df(gen_ambient_load_fractions, gen_mt_fraction_per_cell),
  raw_counts = gen_raw_counts, ambient_counts = gen_raw_ambient, endogenous_counts = gen_raw_endogenous
)
gen_var_by_cellular_params <- make_barplot_of_explained_variance(gen_var_explained)
saveRDS(gen_var_by_cellular_params, here::here("results/main_figures/figure_2/gen_var_by_celular_params.rds"))
gen_var_explained$dataset <- "genotype"
write_xlsx(gen_var_explained, path = here::here("results/supplemental_tables/gen_var_by_cellular_parameters.xlsx"))
rm(gen_var_by_cellular_params)
```

------------------------------------------------------------------------

## Gene-level ambient load vs feature properties

### Background expression profile (empty droplets)

```{r}
lc_background <- calc_background_profile(
  here::here("data/raw/low_complexity/raw_feature_bc_matrix"),
  object = lc_seu,
  threshold = 50
)
lc_background_mouse <- lc_background[grep("^mm10",   names(lc_background))]
lc_background_human <- lc_background[grep("^GRCh38", names(lc_background))]

# Medium complexity: average background across samples
mc_background_human <- Reduce("+", lapply(mc_human_samples, function(samp) {
  calc_background_profile(
    here::here(paste0("data/raw/medium_complexity/", samp, "_STARsolo_Alignment/", samp, "_Solo.out/GeneFull_Ex50pAS/raw/")),
    object = mc_human_seu, threshold = 50
  )
})) / length(mc_human_samples)
mc_background_human_m <- mc_background_human[grep("^mm10",   names(mc_background_human))]
mc_background_human_h <- mc_background_human[grep("^GRCh38", names(mc_background_human))]

mc_background_mouse <- Reduce("+", lapply(mc_mouse_samples, function(samp) {
  calc_background_profile(
    here::here(paste0("data/raw/medium_complexity/", samp, "_STARsolo_Alignment/",
                      samp, "_Solo.out/GeneFull_Ex50pAS/raw/")),
    object = mc_mouse_seu, threshold = 50
  )
})) / length(mc_mouse_samples)
mc_background_mouse_m <- mc_background_mouse[grep("^mm10",   names(mc_background_mouse))]
mc_background_mouse_h <- mc_background_mouse[grep("^GRCh38", names(mc_background_mouse))]

hc_background <- Reduce("+", lapply(hc_samples, function(samp) {
  calc_background_profile(
    here::here(paste0("data/raw/high_complexity/", samp, "_STARsolo_Alignment/",
                      samp, "_Solo.out/GeneFull_Ex50pAS/raw/")),
    object = hc_seu, threshold = 50
  )
})) / length(hc_samples)
hc_background_mouse <- hc_background[grep("^mm10",   names(hc_background))]
hc_background_human <- hc_background[grep("^GRCh38", names(hc_background))]

gen_background <- Reduce("+", lapply(gen_samples, function(samp) {
  calc_background_profile(
    here::here(paste0("data/raw/genotype/h5-files/", samp, "_raw_feature_bc_matrix.h5")),
    object = gen_seu, threshold = 50
  )
})) / length(gen_samples)
```

### Ambient load vs background expression, endogenous UMIs, endogenous features, and endogenous expression frequency

These analyses follow the same pattern for each dataset. Only the core compute + save steps are shown; plotting functions are defined in `02_functions.qmd`.

```{r}
# Helper: save a scatterplot RDS and an xlsx table, then remove the data frame
save_scatter <- function(df, plot_fn, plot_path, table_path, extra_cols = list()) {
  p <- plot_fn(df)
  saveRDS(p, plot_path)
  for (col in names(extra_cols)) df[[col]] <- extra_cols[[col]]
  write_xlsx(df, path = table_path)
  rm(df, p)
}

# --- Ambient load vs total background expression ---
for (ds in list(
  list(name = "lc", df = rbind(
    make_df_of_ambient_load_vs_total_background_expression("human", lc_background_mouse, lc_raw_human_ambient),
    make_df_of_ambient_load_vs_total_background_expression("mouse", lc_background_human, lc_raw_mouse_ambient)
  ), dataset = "low_complexity", fig = "lc"),
  list(name = "mc", df = rbind(
    make_df_of_ambient_load_vs_total_background_expression("human", mc_background_human, mc_raw_human_ambient),
    make_df_of_ambient_load_vs_total_background_expression("mouse", mc_background_mouse, mc_raw_mouse_ambient)
  ), dataset = "medium_complexity", fig = "mc"),
  list(name = "hc", df = rbind(
    make_df_of_ambient_load_vs_total_background_expression("human", hc_background_mouse, hc_raw_human_ambient),
    make_df_of_ambient_load_vs_total_background_expression("mouse", hc_background_human, hc_raw_mouse_ambient)
  ), dataset = "high_complexity", fig = "hc"),
  list(name = "gen", df = data.frame(
    log_background_expression = log1p(gen_background[match(names(rowSums(gen_raw_ambient, na.rm = TRUE)), names(gen_background))]),
    log_ambient_expression    = log1p(rowSums(gen_raw_ambient, na.rm = TRUE))
  ), dataset = "genotype", fig = "gen")
)) {
  save_scatter(
    ds$df,
    make_scatterplot_of_ambient_load_vs_total_background_expression,
    here::here(paste0("results/main_figures/figure_2/", ds$fig, "_amb_load_vs_bcg_exp.rds")),
    here::here(paste0("results/supplemental_tables/amb_load_vs_bckg_expression/", ds$fig, "_amb_load_vs_background_expression.xlsx")),
    list(dataset = ds$dataset)
  )
}

# Summary: mean Pearson R across datasets
# (Computed on non-zero ambient genes)
```

------------------------------------------------------------------------

# Ambient removal and endogenous retention

## Method abbreviations

```{r}
tools <- c("CellBender", "CellClear", "FastCAR", "scAR", "scCDC",
           "SoupX", "SoupX_r", "DecontX", "DecontX_r")
tools_abbr <- c("cb", "cc", "fc", "ar", "cd", "sx", "sxr", "dx", "dxr")
tools_map <- setNames(tools_abbr, tools)
```

## Setup corrected count lists

### Low complexity

```{r}
lc_corrected_counts_human <- divide_ambient_and_endogenous_per_tool(seu = lc_seu, extract = "human")
lc_corrected_counts_mouse <- divide_ambient_and_endogenous_per_tool(seu = lc_seu, extract = "mouse")
```

### Medium complexity

```{r}
mc_corrected_counts_human <- divide_ambient_and_endogenous_per_tool(seu = mc_human_seu, extract = "mouse")
mc_corrected_counts_mouse <- divide_ambient_and_endogenous_per_tool(seu = mc_mouse_seu, extract = "human")

# Combine human and mouse into one list with padded zero matrices
mc_corrected_human <- list()
mc_corrected_mouse <- list()

for (name in names(mc_corrected_counts_mouse)) {
  mouse_mat <- mc_corrected_counts_mouse[[name]]
  rownames(mouse_mat) <- paste0("Mouse_", rownames(mouse_mat))

  human_mat <- mc_corrected_counts_human[[name]]
  rownames(human_mat) <- paste0("Human_", rownames(human_mat))

  # Zero-pad so both matrices share the same rownames
  pad_human <- Matrix::Matrix(0, nrow = nrow(mouse_mat), ncol = ncol(human_mat), sparse = TRUE, dimnames = list(rownames(mouse_mat), colnames(human_mat)))
  pad_mouse <- Matrix::Matrix(0, nrow = nrow(human_mat), ncol = ncol(mouse_mat), sparse = TRUE, dimnames = list(rownames(human_mat), colnames(mouse_mat)))

  human_mat <- Matrix::rbind2(human_mat, pad_human)
  mouse_mat <- Matrix::rbind2(pad_mouse, mouse_mat)

  mc_corrected_human[[name]] <- human_mat
  mc_corrected_mouse[[name]] <- mouse_mat
}
mc_corrected_counts <- mapply(Matrix::cbind2, mc_corrected_mouse, mc_corrected_human, SIMPLIFY = FALSE)
```

### High complexity

```{r}
hc_corrected_counts_human <- divide_ambient_and_endogenous_per_tool(seu = hc_seu, extract = "human")
hc_corrected_counts_mouse <- divide_ambient_and_endogenous_per_tool(seu = hc_seu, extract = "mouse")
```

### Genotype

```{r}
gen_corrected_counts <- list()

endo_snp <- csc_all[csc_all$Ambient == 0, ]
ambi_snp <- csc_all[csc_all$Ambient != 0, ]
rn <- rownames(gen_raw_counts)
cn <- colnames(gen_raw_counts)

gen_corrected_counts$input_endogenous <- make_sparse_subset(endo_snp$Symbol, endo_snp$Barcode, rn, cn, gen_raw_counts)
gen_corrected_counts$input_ambient <- make_sparse_subset(ambi_snp$Symbol, ambi_snp$Barcode, rn, cn, gen_raw_counts)

assay_abbrev <- c(
  CellBender = "cb", SoupXfull = "sx", SoupXreduced = "sxr",
  DecontXfull = "dx", DecontXreduced = "dxr",
  FastCAR = "fc", scCDC = "cd", scAR = "ar", CellClear = "cc"
)
for (assay in names(assay_abbrev)) {
  abbr <- assay_abbrev[[assay]]
  mat <- as(gen_seu@assays[[assay]]$counts, "dgCMatrix")
  gen_corrected_counts[[paste0(abbr, "_endogenous")]] <- make_sparse_subset(endo_snp$Symbol, endo_snp$Barcode, rn, cn, mat)
  gen_corrected_counts[[paste0(abbr, "_ambient")]] <- make_sparse_subset(ambi_snp$Symbol, ambi_snp$Barcode, rn, cn, mat)
}
```

### Smart-seq2

```{r}
smart_corrected_counts <- get_total_counts_per_tool(seu = smart_seu)
```

### Synthetic

The synthetic dataset uses the real ambient matrix from the simulations to define ambient ground truth (genes present in real_ambient = ambient; genes absent = endogenous).

```{r}
synt_simulations <- list()
for (sample in synt_samples) {
  synt_simulations[[sample]] <- readRDS(here::here(paste0(
    "data/raw/synthetic/dataset_R_objects_full/", sample, ".Rds"
  )))
}

# Convert Ensembl IDs to gene symbols for the ambient matrices
synt_sims <- lapply(synt_samples, function(sample) {
  mat <- convert_ensembl_ids_to_gene_names(synt_simulations[[sample]]$real_ambient)
  colnames(mat) <- paste(colnames(mat), sample, sep = "_")
  mat
})
names(synt_sims) <- synt_samples

# Genes absent from real_ambient → endogenous; genes present → ambient
df_endo <- do.call(rbind, lapply(names(synt_sims), function(sample_name) {
  mat <- synt_sims[[sample_name]]
  zero_genes <- rownames(mat)[Matrix::rowSums(mat) == 0]
  expand.grid(feature = zero_genes, barcode = colnames(mat),
  EEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE)
}))

df_amb <- do.call(rbind, lapply(names(synt_sims), function(sample_name) {
  mat <- synt_sims[[sample_name]]
  non_zero_genes <- rownames(mat)[Matrix::rowSums(mat) != 0]
  expand.grid(feature = non_zero_genes, barcode = colnames(mat),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE)
}))

synt_raw_counts <- synt_seu@assays$RNA$counts
rn_synt <- rownames(synt_raw_counts)
cn_synt <- colnames(synt_raw_counts)

synt_corrected_counts <- list()
synt_corrected_counts$input_endogenous <- make_sparse_subset(df_endo$feature, df_endo$barcode, rn_synt, cn_synt, synt_raw_counts)
synt_corrected_counts$input_ambient <- make_sparse_subset(df_amb$feature,  df_amb$barcode,  rn_synt, cn_synt, synt_raw_counts)

synt_raw_counts <- as(synt_raw_counts, "dgCMatrix")

for (assay in names(assay_abbrev)) {
  abbr <- assay_abbrev[[assay]]
  corrected <- as(synt_seu@assays[[assay]]$counts, "dgCMatrix")
  synt_corrected_counts[[paste0(abbr, "_endogenous")]] <- make_sparse_subset(df_endo$feature, df_endo$barcode, rn_synt, cn_synt, corrected)
  synt_corrected_counts[[paste0(abbr, "_ambient")]] <- make_ambient_subset(df_amb$feature, df_amb$barcode, rn_synt, cn_synt, synt_raw_counts, corrected)
}
```

### Synthetic — cell type division (MCF12A, HCC1500, HS578T)

```{r}
build_synt_celltype_matrices <- function(cell_type, field, seu_obj, simulations) {
  # Raw
  raw <- extract_raw_counts(seu_obj, origin = cell_type)

  # Build cross-sample matrix from simulations
  mat_list <- lapply(seq_along(simulations), function(i) {
    m <- simulations[[i]][[field]]
    colnames(m) <- paste0(colnames(m), "_sample", i)
    m
  })
  combined <- do.call(cbind, mat_list)
  combined <- convert_ensembl_ids_to_gene_names(combined)

  common_rows <- intersect(rownames(combined), rownames(raw))
  common_cols <- intersect(colnames(combined), colnames(raw))
  list(raw = raw, signal = combined[common_rows, common_cols, drop = FALSE])
}

MCF12A_ambient <- build_synt_celltype_matrices("MCF12A",  "real_ambient", synt_seu, synt_simulations)
HCC1500_ambient <- build_synt_celltype_matrices("HCC1500", "real_ambient", synt_seu, synt_simulations)
HS578T_ambient <- build_synt_celltype_matrices("HS578T",  "real_ambient", synt_seu, synt_simulations)

MCF12A_endogenous <- build_synt_celltype_matrices("MCF12A",  "real_counts", synt_seu, synt_simulations)
HCC1500_endogenous <- build_synt_celltype_matrices("HCC1500", "real_counts", synt_seu, synt_simulations)
HS578T_endogenous <- build_synt_celltype_matrices("HS578T",  "real_counts", synt_seu, synt_simulations)
```

## Top-20 most ambient genes per dataset

```{r}
# Low complexity
top20_lc_human <- names(sort(Matrix::rowSums(lc_corrected_counts_human$input_ambient), decreasing = TRUE)[1:20])
top20_lc_mouse <- names(sort(Matrix::rowSums(lc_corrected_counts_mouse$input_ambient), decreasing = TRUE)[1:20])

lc_top20_human <- lapply(lc_corrected_counts_human, function(mat) mat[rownames(mat) %in% top20_lc_human, ])
lc_top20_mouse <- lapply(lc_corrected_counts_mouse, function(mat) mat[rownames(mat) %in% top20_lc_mouse, ])

saveRDS(lc_top20_human, here::here("results/top20_objects/lc_top20_human.rds"))
saveRDS(lc_top20_mouse, here::here("results/top20_objects/lc_top20_mouse.rds"))

# Medium complexity
top20_mc_human <- names(sort(Matrix::rowSums(mc_corrected_human$input_ambient), decreasing = TRUE)[1:20])
top20_mc_mouse <- names(sort(Matrix::rowSums(mc_corrected_mouse$input_ambient), decreasing = TRUE)[1:20])
top20_mc <- c(top20_mc_human, top20_mc_mouse)

mc_top20 <- lapply(mc_corrected_counts, function(mat) mat[rownames(mat) %in% top20_mc, ])
saveRDS(mc_top20, here::here("results/top20_objects/mc_top20.rds"))

# High complexity
top20_hc_human <- names(sort(Matrix::rowSums(hc_corrected_counts_human$input_ambient), decreasing = TRUE)[1:20])
top20_hc_mouse <- names(sort(Matrix::rowSums(hc_corrected_counts_mouse$input_ambient), decreasing = TRUE)[1:20])

hc_top20_human <- lapply(hc_corrected_counts_human, function(mat) mat[rownames(mat) %in% top20_hc_human, ])
hc_top20_mouse <- lapply(hc_corrected_counts_mouse, function(mat) mat[rownames(mat) %in% top20_hc_mouse, ])

saveRDS(hc_top20_human, here::here("results/top20_objects/hc_top20_human.rds"))
saveRDS(hc_top20_mouse, here::here("results/top20_objects/hc_top20_mouse.rds"))

# Genotype
top20_gen <- names(sort(Matrix::rowSums(gen_corrected_counts$input_ambient), decreasing = TRUE)[1:20])
gen_top20 <- lapply(gen_corrected_counts, function(mat) mat[rownames(mat) %in% top20_gen, ])
saveRDS(gen_top20, here::here("results/top20_objects/gen_top20.rds"))
```

## Compartment removal plots and tables

`run_compartment_removal_analysis()` (defined in `02_functions.qmd`) handles plot generation, saving, and table export for both ambient and endogenous compartments.

```{r}
# Low complexity
lc_ambient_results <- run_compartment_removal_analysis(lc_seu, "low_complexity", "ambient",
  here::here("results/supplemental_figures/supp_fig_3/low_complexity"),
  here::here("results/supplemental_tables/ambient_removal/lc_ambient_removal.xlsx"))

lc_endogenous_results <- run_compartment_removal_analysis(lc_seu, "low_complexity", "endogenous",
  here::here("results/supplemental_figures/supp_fig_4/low_complexity"),
  here::here("results/supplemental_tables/endogenous_removal/lc_endogenous_removal.xlsx"))

# High complexity
hc_ambient_results <- run_compartment_removal_analysis(hc_seu, "high_complexity", "ambient",
  here::here("results/supplemental_figures/supp_fig_3/high_complexity"),
  here::here("results/supplemental_tables/ambient_removal/hc_ambient_removal.xlsx"))

hc_endogenous_results <- run_compartment_removal_analysis(hc_seu, "high_complexity", "endogenous",
  here::here("results/supplemental_figures/supp_fig_4/high_complexity"),
  here::here("results/supplemental_tables/endogenous_removal/hc_endogenous_removal.xlsx"))
```

Medium complexity, genotype, smart-seq2, and synthetic removal plots follow the same pattern but require per-species data frames (see mc_corrected_human / mc_corrected_mouse) or per-tool loops. Figures and tables are saved to the corresponding subdirectories under `results/`.

------------------------------------------------------------------------

## Endogenous removal vs total endogenous UMIs per feature

```{r}
# Low complexity
lc_endo_removal_human <- calc_endogenous_removal_vs_total_endogenous_UMIs_per_feature(
  metadata = lc_seu@meta.data, dataset = "low", corrected_counts = lc_corrected_counts_human)
lc_endo_removal_mouse <- calc_endogenous_removal_vs_total_endogenous_UMIs_per_feature(
  metadata = lc_seu@meta.data, dataset = "low", corrected_counts = lc_corrected_counts_mouse)
lc_endogenous_removal_combined <- data.frame(
  method = lc_endo_removal_human[[1]],
  group = lc_endo_removal_human[[2]],
  error = (lc_endo_removal_human$error + lc_endo_removal_mouse$error) / 2,
  fraction_changed = (lc_endo_removal_human$fraction_changed + lc_endo_removal_mouse$fraction_changed) / 2,
  dataset = lc_endo_removal_human[[5]]
)

mc_metadata <- rbind(mc_mouse_seu@meta.data[, c("orig.ident", "nCount_RNA")],
                     mc_human_seu@meta.data[, c("orig.ident", "nCount_RNA")])
mc_endogenous_removal <- calc_endogenous_removal_vs_total_endogenous_UMIs_per_feature(
  metadata = mc_metadata, dataset = "medium", corrected_counts = mc_corrected_counts)

hc_endo_removal_human <- calc_endogenous_removal_vs_total_endogenous_UMIs_per_feature(
  metadata = hc_seu@meta.data, dataset = "high", corrected_counts = hc_corrected_counts_human)
hc_endo_removal_mouse <- calc_endogenous_removal_vs_total_endogenous_UMIs_per_feature(
  metadata = hc_seu@meta.data, dataset = "high", corrected_counts = hc_corrected_counts_mouse)
hc_endogenous_removal_combined <- data.frame(
  method = hc_endo_removal_human[[1]],
  group = hc_endo_removal_human[[2]],
  error = (hc_endo_removal_human$error + hc_endo_removal_mouse$error) / 2,
  fraction_changed  = (hc_endo_removal_human$fraction_changed + hc_endo_removal_mouse$fraction_changed) / 2,
  dataset = hc_endo_removal_human[[5]]
)

gen_endogenous_removal  <- calc_endogenous_removal_vs_total_endogenous_UMIs_per_feature(
  gen_seu@meta.data, "genotype",  gen_corrected_counts)
smart_endogenous_removal <- calc_endogenous_removal_vs_total_endogenous_UMIs_per_feature(
  smart_seu@meta.data, "smart-seq2", smart_corrected_counts)
synt_endogenous_removal <- calc_endogenous_removal_vs_total_endogenous_UMIs_per_feature(
  synt_seu@meta.data,  "synthetic",  synt_corrected_counts)
synt_wo_ambient_endogenous_removal <- calc_endogenous_removal_vs_total_endogenous_UMIs_per_feature(
  synt_wo_ambient_seu@meta.data, "synt_wo_ambient", synt_wo_ambient_corrected_counts)
```

## Ambient removal vs total ambient UMIs per feature

```{r}
lc_amb_removal_human <- calc_ambient_removal_vs_total_ambient_UMIs_per_feature(
  lc_seu@meta.data, "low", lc_corrected_counts_human)
lc_amb_removal_mouse <- calc_ambient_removal_vs_total_ambient_UMIs_per_feature(
  lc_seu@meta.data, "low", lc_corrected_counts_mouse)
lc_ambient_removal_combined <- data.frame(
  method = lc_amb_removal_human[[1]],
  group = lc_amb_removal_human[[2]],
  error = (lc_amb_removal_human$error + lc_amb_removal_mouse$error) / 2,
  fraction_changed = (lc_amb_removal_human$fraction_changed + lc_amb_removal_mouse$fraction_changed) / 2,
  dataset = lc_amb_removal_human[[5]]
)

mc_corrected_counts_combined <- Matrix::cbind2(
  do.call(Matrix::cbind2, mc_corrected_mouse),
  do.call(Matrix::cbind2, mc_corrected_human)
)
mc_ambient_removal <- calc_ambient_removal_vs_total_ambient_UMIs_per_feature(
  mc_metadata, "medium", mc_corrected_counts)

hc_amb_removal_human <- calc_ambient_removal_vs_total_ambient_UMIs_per_feature(
  hc_seu@meta.data, "high", hc_corrected_counts_human)
hc_amb_removal_mouse <- calc_ambient_removal_vs_total_ambient_UMIs_per_feature(
  hc_seu@meta.data, "high", hc_corrected_counts_mouse)
hc_ambient_removal_combined <- data.frame(
  method = hc_amb_removal_human[[1]],
  group = hc_amb_removal_human[[2]],
  error = (hc_amb_removal_human$error + hc_amb_removal_mouse$error) / 2,
  fraction_changed = (hc_amb_removal_human$fraction_changed + hc_amb_removal_mouse$fraction_changed) / 2,
  dataset = hc_amb_removal_human[[5]]
)

gen_ambient_removal <- calc_ambient_removal_vs_total_ambient_UMIs_per_feature(
  gen_seu@meta.data, "genotype",  gen_corrected_counts)
synt_ambient_removal <- calc_ambient_removal_vs_total_ambient_UMIs_per_feature(
  synt_seu@meta.data, "synthetic", synt_corrected_counts)
```

## Summary statistics: removal fractions

```{r}
for (method in tools) {
  # Ambient removal
  amb_df <- ambient_removal[ambient_removal$decontamination_method == method, ]
  amb_df$fraction_ambient_removed <- as.numeric(as.character(amb_df$fraction_ambient_removed))
  mean_amb <- mean(amb_df$fraction_ambient_removed[
    amb_df$fraction_ambient_removed >= 0 & is.finite(amb_df$fraction_ambient_removed)
  ])

  # Endogenous removal
  endo_df <- endogenous_removal[endogenous_removal$decontamination_method == method, ]
  endo_df$fraction_endogenous_removed <- as.numeric(as.character(endo_df$fraction_endogenous_removed))
  mean_endo <- mean(endo_df$fraction_endogenous_removed[
    endo_df$fraction_endogenous_removed >= 0 & is.finite(endo_df$fraction_endogenous_removed)
  ])

  message(method, " — mean ambient removed: ", round(mean_amb, 3),
          " | mean endogenous removed: ", round(mean_endo, 3))
}
```

## Combined removal summary (dot plot)

```{r}
# Build per-dataset removal data frames
make_removal_df <- function(amb, endo, smart_endo = NULL, dataset_label) {
  if (!is.null(smart_endo)) {
    colnames(smart_endo)[3:4] <- c("error_endogenous", "fraction_changed_endogenous")
    smart_endo$dataset <- dataset_label
    return(smart_endo)
  }
  data.frame(
    method = amb[[1]],
    group = amb[[2]],
    error_endogenous = endo$error,
    error_ambient = amb$error,
    fraction_changed_endogenous = endo$fraction_changed,
    fraction_changed_ambient = amb$fraction_changed,
    dataset = dataset_label
  )
}

lc_removal_combined <- make_removal_df(lc_ambient_removal_combined,   lc_endogenous_removal_combined, dataset_label = "low")
mc_removal_combined <- make_removal_df(mc_ambient_removal,            mc_endogenous_removal, dataset_label = "medium")
hc_removal_combined <- make_removal_df(hc_ambient_removal_combined,   hc_endogenous_removal_combined, dataset_label = "high")
gen_removal_combined <- make_removal_df(gen_ambient_removal,           gen_endogenous_removal, dataset_label = "genotype")
synt_removal_combined <- make_removal_df(synt_ambient_removal,          synt_endogenous_removal, dataset_label = "synthetic")

saveRDS(lc_removal_combined, here::here("results/main_figures/figure_4/lc_removal_combined.Rds"))
saveRDS(mc_removal_combined, here::here("results/main_figures/figure_4/mc_removal_combined.Rds"))
saveRDS(hc_removal_combined, here::here("results/main_figures/figure_4/hc_removal_combined.Rds"))
saveRDS(gen_removal_combined, here::here("results/main_figures/figure_4/gen_removal_combined.Rds"))
saveRDS(synt_removal_combined, here::here("results/main_figures/figure_4/synt_removal_combined.Rds"))

smart_removal_endogenous <- smart_endogenous_removal
colnames(smart_removal_endogenous)[3:4] <- c("error_endogenous", "fraction_changed_endogenous")
saveRDS(smart_removal_endogenous, here::here("results/main_figures/figure_4/smart_removal_endogenous.Rds"))

synt_wo_ambient_removal_endogenous <- synt_wo_ambient_endogenous_removal
colnames(synt_wo_ambient_removal_endogenous)[3:4] <- c("error_endogenous", "fraction_changed_endogenous")
saveRDS(synt_wo_ambient_removal_endogenous, here::here("results/main_figures/figure_4/synt_wo_ambient_removal_endogenous.Rds"))

# Combine all datasets for dot plot
removal_all <- dplyr::bind_rows(
  lc_removal_combined, mc_removal_combined, hc_removal_combined,
  gen_removal_combined, smart_removal_endogenous,
  synt_wo_ambient_removal_endogenous, synt_removal_combined
)
removal_all$dataset <- factor(removal_all$dataset,
  levels = c("low", "medium", "high", "genotype", "synthetic", "synt_wo_ambient", "smart-seq2"))

saveRDS(removal_all, here::here("results/main_figures/figure_4/removal_all.Rds"))

dotplot <- ggplot(removal_all, aes(x = dataset, y = method,
                                    color = error_ambient, size = error_endogenous)) +
  geom_point(stroke = 0) +
  scale_color_viridis_c(option = "plasma", limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  plot_theme() +
  labs(x = "Dataset")

corr_plot <- ggplot(removal_all, aes(error_ambient, error_endogenous, color = method)) +
  geom_point() +
  plot_theme() +
  scale_color_manual(values = custom_colors)
```

------------------------------------------------------------------------

# Biology

## Synthetic dataset: evaluate sample 3

```{r}
marker_list <- defineCleanMarkers(synt_wo_ambient_seu)

unclean <- subset(synt_seu, orig.ident == "sample3"); unclean$type <- "unclean"
clean <- subset(synt_wo_ambient_seu, orig.ident == "sample3"); clean$type <- "clean"

synthetic_evaluation <- evaluateSynthetic(
  list(clean = clean, unclean = unclean),
  marker_list = marker_list
)

data <- synthetic_evaluation[[1]]
data$Method <- factor(data$Method, levels = c(
  "RNA", "CellBender", "scAR", "DecontXfull", "DecontXreduced",
  "SoupXfull", "SoupXreduced", "CellClear", "scCDC", "FastCAR"
))
data$deltaModule <- data$Module - data$deltaModule

write_xlsx(data, path = here::here("results/supplemental_tables/synthetic_results.xlsx"))

for (metric in c("ARI", "Distance", "Module", "deltaModule")) {
  p <- ggplot(data, aes(x = Method, y = .data[[metric]], fill = Type)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_manual(values = custom_colors) +
    plot_theme()
  ggsave(here::here(paste0("results/main_figures/figure_4/BioInterpret_Synthetic_", metric, ".pdf")), plot = p, width = 10, height = 10)
}

pdf(here::here("results/main_figures/figure_4/BioInterpret_Synthetic.pdf"), width = 50, height = 40, useDingbats = FALSE)
print(patchwork::wrap_plots(synthetic_evaluation[[2]], ncol = 5, nrow = 4, byrow = TRUE))
dev.off()
```

## Integration quality

```{r}
datasets_integration <- list(
  Synthetic = list(Path = here::here("data/processed/synthetic/seurat_object/synt_seu.Rds"),
  markers = c("KRT19", "ANKRD30A", "RAB11FIP1", "COL1A2", "VIM", "POSTN", "S100A2", "RPL11", "RPS27")),
  Synthetic_Clean = list(Path = here::here("data/processed/synthetic/seurat_object/synt_seu_wo_ambient.Rds"),
  markers = c("KRT19", "ANKRD30A", "RAB11FIP1", "COL1A2", "VIM", "POSTN", "S100A2", "RPL11", "RPS27")),
  Medium_Mouse = list(Path = here::here("data/processed/medium_complexity/seurat_object/mc_mouse_seu.Rds"),
  markers = c("mm10-Gcg", "mm10-Ttr", "mm10-Irx2", "mm10-Ins2", "mm10-Iapp", "mm10-G6pc2", "mm10-Ppy", "mm10-Id2", "mm10-Gcnt3", "mm10-Sst", "mm10-Rbp4", "mm10-Lepr")),
  Medium_Human = list(Path = here::here("data/processed/medium_complexity/seurat_object/mc_human_seu.Rds"),
  markers = c("GRCh38-GCG", "GRCh38-TTR", "GRCh38-IRX2", "GRCh38-INS", "GRCh38-IAPP", "GRCh38-G6PC2", "GRCh38-PPY", "GRCh38-ID2", "GRCh38-GCNT3", "GRCh38-SST", "GRCh38-RBP4", "GRCh38-LEPR")),
  High = list(Path = here::here("data/processed/high_complexity/seurat_object/hc_seu.Rds"),
  markers = c("GRCh38-GCG", "GRCh38-TTR", "GRCh38-IRX2", "GRCh38-INS", "GRCh38-IAPP", "GRCh38-G6PC2", "mm10-Emcn", "mm10-Nphs2", "mm10-Slc5a2", "mm10-Slc12a3", "mm10-C1qb", "mm10-Siglech")),
  Genotype = list(Path = here::here("data/processed/genotype/seurat_object/gen_seu.Rds"),
  markers = c("Emcn", "Nphs2", "Slc5a2", "Slc12a3", "C1qb", "Siglech", "Igkc", "Nkg7", "S100a4"))
)

integration <- plyr::llply(datasets_integration, .progress = "text", .fun = function(x) {
  out_file <- gsub(".Rds", "_integration.Rds", x$Path)
  if (file.exists(out_file)) return(readRDS(out_file))
  object <- readRDS(x$Path)
  result <- tryCatch(evaluateIntegration(object, known_markers = x$markers), error = function(e) "Failed")
  if (is.data.frame(result)) saveRDS(result, out_file)
  rm(object); gc()
  result
})
```

## Per-sample biology

```{r}
datasets_biology <- list(
  Low = list(Path = here::here("data/processed/low_complexity/seurat_object/lc_seu.Rds"),
  markers = c("GRCh38-POLR2J3", "GRCh38-RWDD1", "GRCh38-DUXAP8", "mm10-Atxn7l3b", "mm10-Son", "mm10-Sec62"),
  evalNhoods = TRUE,  evalDeNovo = TRUE,  species_label = "origin"),
  Medium_Mouse  = list(Path = here::here("data/processed/medium_complexity/seurat_object/mc_mouse_seu.Rds"),
  markers = c("mm10-Gcg", "mm10-Ttr", "mm10-Irx2", "mm10-Ins2", "mm10-Iapp", "mm10-G6pc2"),
  evalNhoods = TRUE,  evalDeNovo = TRUE,  species_label = "origin"),
  Medium_Human = list(Path = here::here("data/processed/medium_complexity/seurat_object/mc_human_seu.Rds"),
  markers = c("GRCh38-GCG", "GRCh38-TTR", "GRCh38-IRX2", "GRCh38-INS", "GRCh38-IAPP", "GRCh38-G6PC2"),
  evalNhoods = TRUE,  evalDeNovo = TRUE,  species_label = "origin"),
  High = list(Path = here::here("data/processed/high_complexity/seurat_object/hc_seu.Rds"),
  markers = c("GRCh38-GCG", "GRCh38-INS", "mm10-Emcn", "mm10-Nphs2", "mm10-C1qb"),
                       evalNhoods = TRUE,  evalDeNovo = TRUE,  species_label = "origin"),
  Genotype = list(Path = here::here("data/processed/genotype/seurat_object/gen_seu.Rds"),
  markers = c("Emcn", "Nphs2", "Slc5a2", "Slc12a3", "C1qb"),
  evalNhoods = FALSE, evalDeNovo = TRUE,  species_label = "origin"),
  Synthetic = list(Path = here::here("data/processed/synthetic/seurat_object/synt_seu.Rds"),
  markers = c("KRT19", "ANKRD30A", "RAB11FIP1", "COL1A2", "VIM", "POSTN"),
  evalNhoods = TRUE,  evalDeNovo = FALSE, species_label = "celltype"),
  Synthetic_Clean = list(Path = here::here("data/processed/synthetic/seurat_object/synt_seu_wo_ambient.Rds"),
  markers = c("KRT19", "ANKRD30A", "RAB11FIP1", "COL1A2", "VIM", "POSTN"),
  evalNhoods = TRUE,  evalDeNovo = FALSE, species_label = "celltype")
)

sample_biology <- plyr::llply(datasets_biology, .progress = "text", .fun = function(x) {
  out_file <- gsub(".Rds", "_sample_biology.Rds", x$Path)
  if (file.exists(out_file)) return(readRDS(out_file))
  object <- readRDS(x$Path)
  result <- tryCatch(
    evaluatePerSampleBiology(object, known_markers = x$markers,
                             species_label = x$species_label,
                             evalNhoods = x$evalNhoods,
                             evalDeNovo = x$evalDeNovo),
    error = function(e) "Failed"
  )
  if (is.data.frame(result)) saveRDS(result, out_file)
  rm(object); gc()
  result
})
```

## Label transfer

Label transfer is performed for medium complexity (pancreas reference), high complexity (pancreas reference), and genotype (kidney reference). The synthetic dataset is excluded from label transfer evaluation.

```{r}
datasets_labels <- list(
  Medium_Mouse = list(Path = here::here("data/processed/medium_complexity/seurat_object/mc_mouse_seu.Rds"), reference_path = "Pancreas", subsetTo = "mouse",  split_genes_by_origin = TRUE),
  Medium_Human = list(Path = here::here("data/processed/medium_complexity/seurat_object/mc_human_seu.Rds"), reference_path = "Pancreas", subsetTo = "human",  split_genes_by_origin = TRUE),
  High = list(Path = here::here("data/processed/high_complexity/seurat_object/hc_seu.Rds"), reference_path = "Pancreas", subsetTo = "human",  split_genes_by_origin = TRUE),
  Genotype = list(Path = here::here("data/processed/genotype/seurat_object/gen_seu.Rds"), reference_path = "Kidney",   subsetTo = NULL,     split_genes_by_origin = FALSE)
)

labels <- plyr::llply(datasets_labels, .progress = "text", .fun = function(x) {
  out_file <- gsub(".Rds", "_labels.Rds", x$Path)
  if (file.exists(out_file)) return(readRDS(out_file))
  object <- readRDS(x$Path)
  result <- tryCatch(
    evaluateLabelTransfer(object, reference_path = x$reference_path, subsetTo = x$subsetTo, split_genes_by_origin = x$split_genes_by_origin),
    error = function(e) "Failed"
  )
  if (is.data.frame(result)) saveRDS(result, out_file)
  rm(object); gc()
  result
})
```

## Collect results and save figures

```{r}
labels_res <- collectResults(labels, columns = c("Mapping", "Annotation", "Purity", "RMSE", "cASW", "ARI"))
ggsave(here::here("results/main_figures/figure_5/labels_results.pdf"),
       plot = labels_res[[1]], width = 10, height = 10)
write_xlsx(labels_res[[2]], path = here::here("results/main_figures/figure_5/labels_results.xlsx"))

integrated_res <- collectResults(integration, columns = c("LISI", "ASW", "PCR", "ARI", "Gini", "Entropy", "Morans"), subsetColumn = "State", subsetTo = "Integrated")
unintegrated_res <- collectResults(integration, columns = c("LISI", "ASW", "PCR", "ARI", "Gini", "Entropy", "Morans"), subsetColumn = "State", subsetTo = "Unintegrated")
ggsave(here::here("results/main_figures/figure_5/integrated_results.pdf"),
       plot = integrated_res[[1]], width = 10, height = 10)
ggsave(here::here("results/main_figures/figure_5/unintegrated_results.pdf"),
       plot = unintegrated_res[[1]], width = 10, height = 10)
write_xlsx(list(integrated = integrated_res[[2]], unintegrated = unintegrated_res[[2]]), path = here::here("results/main_figures/figure_5/integration_results.xlsx"))

samples_res <- collectResults(sample_biology, columns = c("Purity", "Species", "Markers", "Entropy", "Morans", "Gini"))
ggsave(here::here("results/main_figures/figure_5/sample_biology_results.pdf"),
       plot = samples_res[[1]], width = 10, height = 10)
write_xlsx(samples_res[[2]], path = here::here("results/main_figures/figure_5/sample_biology_results.xlsx"))
```
